---
title: "Week 3: Generalized Linear Models"
subtitle: "PS 818 - Data Analysis with Statistical Models"
author:
  - name: "Anton Strezhnev"
    affiliation: "University of Wisconsin-Madison"
    email: strezhnev@wisc.edu
date: "September 21, 2026"
date-format: "MMMM D, YYYY"
editor: 
  render-on-save: true
format: 
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    overview: true
    navigation-mode: vertical
    progress: false
    controls: true
    link-external-newwindow: true
    footer: "PS 818 - University of Wisconsin-Madison"
    html-math-method:
      method: mathjax
      url: https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js
execute: 
  cache: false
embed-resources: true
bibliography: ../ps818.bib
link-citations: true

---

## Generalized Linear Models {.title-slide}

::: {style="display: none"}
$$
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}
$$
:::

```{r, echo = F, message=F, warnings=F}
library(tidyverse)
library(knitr)
library(broom)
library(estimatr)
options(digits=3)
```

## Previously

::: {.incremental}

- **Likelihood**: a function of the parameters $\theta$ evaluated at the observed data

  $$\mathcal{L}(\theta| \mathbf{y}) = f(\mathbf{y}| \theta) \qquad \ell_n(\theta) = \sum_{i=1}^n \log f(y_i|\theta)$$

- **Maximum likelihood estimator**: $\hat{\theta} = \argmax_{\theta \in \Theta} \ \ell_n(\theta)$
  - **Consistent**: $\hat{\theta} \overset{p}{\to} \theta_0$
  - **Asymptotically normal**: $\sqrt{n}(\hat{\theta} - \theta_0) \overset{d}{\to} \text{Normal}\big(0, \mathcal{I}(\theta_0)^{-1}\big)$
  - **Asymptotically efficient**: attains the Cramér-Rao lower bound

:::

## This week

::: {.incremental}

- **More properties of the MLE**
  - **Invariance** and the **delta method**
- **How does numerical optimization actually work?**
  - **Newton-Raphson**, Fisher scoring, gradient ascent
- **Generalized linear models**
  - Extending the classic linear form to **functions** of the CEF
  - $g(\mathbb{E}[Y_i|X_i]) = X_i^{\prime}\beta$
- **Binary outcome models**
  - Logit and probit; latent-variable formulations
  - Ordinal and multinomial extensions

:::

## More on the MLE

## Invariance

::: {.incremental}

- We often want to do inference on some **function** of $\theta$
  - (e.g.) we estimate $\beta$ but want a predicted probability, a risk ratio, or a marginal effect.
- The MLE has a convenient property: **invariance**.
  - If $\hat{\theta}$ is the MLE of $\theta$, then $g(\hat{\theta})$ is the MLE of $g(\theta)$ -- for *any* function $g$.
- Combined with consistency, $g(\hat{\theta}) \overset{p}{\to} g(\theta_0)$ for continuous $g$ (continuous mapping theorem).
- But how do we get the standard errors (asymptotically)?
  - The **delta method**

:::

## Delta method

::: {.incremental}

- We know the asymptotic variance of $\hat{\theta}$
  - It's the inverse Fisher information, estimated by the inverse negative Hessian at the MLE.
- To get $\text{Var}(g(\hat{\theta}))$, start with a first-order Taylor approximation around the true value $\theta_0$

  $$g(\hat{\theta}) \approx g(\theta_0) + \big[\nabla g(\theta_0)^{\prime}\big](\hat{\theta} - \theta_0)$$

- Take the variance

  $$\text{Var}\big(g(\hat{\theta})\big) \approx \text{Var}\bigg(g(\theta_0) + \big[\nabla g(\theta_0)^{\prime}\big]\hat{\theta} - \big[\nabla g(\theta_0)^{\prime}\big]\theta_0\bigg)$$

- $g(\theta_0)$ and $\nabla g(\theta_0)$ are **constants**, so the variance of the first and third terms is zero

  $$\text{Var}\big(g(\hat{\theta})\big) \approx \text{Var}\bigg(\big[\nabla g(\theta_0)^{\prime}\big]\hat{\theta}\bigg)$$

:::

## Delta method

::: {.incremental}

- Pull out and "square" the constant. We get an expression in terms of the original variance-covariance matrix

  $$\text{Var}\big(g(\hat{\theta})\big) \approx \big[\nabla g(\theta_0)^{\prime}\big]\text{Var}(\hat{\theta})\big[\nabla g(\theta_0)\big]$$

- And so our asymptotic approximation is

  $$\sqrt{n}\big(g(\hat{\theta}) - g(\theta_0)\big) \overset{d}{\to} \text{Normal}\bigg(0, \ \big[\nabla g(\theta_0)^{\prime}\big]\mathcal{I}(\theta_0)^{-1}\big[\nabla g(\theta_0)\big]\bigg)$$

- We can use a plug-in estimator at the MLE: evaluate the gradient at $\hat{\theta}$ and use our existing estimator $\widehat{\text{Var}}(\hat{\theta})$.

:::

## How optimizers work

## Finding the maximum

::: {.incremental}

- How does `maxLik()` actually find a solution?
- Our estimator is defined as

  $$\hat{\theta} = \argmax_{\theta \in \Theta} \ \ell_n(\theta)$$

- For well-behaved problems the log-likelihood is **concave**, so finding the maximum amounts to finding the unique *solution* to $S_n(\theta) = 0$.
- Closed form solutions are rare (the Normal linear model is a special case).
  - Generally, we'll use an *iterative* algorithm to find this solution.

:::

## Newton-Raphson

::: {.incremental}

- We start at some initial guess $\hat{\theta}^{(0)}$
- Let $\hat{\theta}^{(t)}$ denote the "current" value and $\hat{\theta}^{(t+1)}$ our update -- we keep iterating until convergence.
- Our goal is to solve for a zero of $S_n(\theta)$. Take a first-order Taylor approximation around our current guess

  $$S_n(\theta) \approx S_n(\hat{\theta}^{(t)}) + \nabla S_n(\hat{\theta}^{(t)}) \left(\theta - \hat{\theta}^{(t)}\right)$$

- What's $\nabla S_n(\hat{\theta}^{(t)})$?
  - It's the Jacobian of the gradient... or the matrix of second-order partial derivatives of the log-likelihood... or the **Hessian** $H_n(\hat{\theta}^{(t)})$!

:::

## Newton-Raphson

::: {.incremental}

- Our next value of $\hat{\theta}$ is the value that sets this approximation to zero

  $$0 =  S_n(\hat{\theta}^{(t)}) + H_n(\hat{\theta}^{(t)}) \left(\hat{\theta}^{(t+1)} - \hat{\theta}^{(t)}\right)$$

- Multiply through by the inverse Hessian and rearrange

  $$\hat{\theta}^{(t+1)} =  \hat{\theta}^{(t)} - H_n^{-1}(\hat{\theta}^{(t)}) S_n(\hat{\theta}^{(t)})$$

- Recall that the **negative** Hessian at $\hat\theta$ is the **observed Fisher information** $\mathcal{J}_n(\hat\theta) = -H_n(\hat\theta)$
  - So the update is the score scaled by the inverse negative hessian.
- An alternative algorithm, **Fisher scoring**, substitutes the *expected* Fisher information $\mathcal{I}_n$ for the observed one.
  - Identical for GLMs with the canonical link -- we'll see why later.

:::

## Gradient ascent

::: {.incremental}

- Compare the Newton-Raphson update to the classic "gradient descent/gradient ascent" algorithm

  $$\hat{\theta}^{(t+1)} =  \hat{\theta}^{(t)} + \gamma_t S_n(\hat{\theta}^{(t)})$$

- Gradient ascent chooses some sequence of $\gamma_t$ (the "learning rate" or "step size") at iteration $t$
  - The gradient tells us **which direction** to move in and $\gamma_t$ tells us **how far** to move.
- Contrast with Newton-Raphson, where "how far to move" comes from computing and inverting the Hessian
  - More information per step, but each step is computationally expensive (inverting a matrix is computationally hard)
- **Trade-off**: Newton-Raphson converges quickly but each step is costly; gradient ascent steps are cheap but need more of them.
  - Also typically need to **tune** the learning rate $\gamma_t$ - too large and we **overshoot** the true optimum

:::


## Illustration

::: {.incremental}

- Let's illustrate with a simple one-parameter likelihood.

  $$y_i \underset{\text{i.i.d}}{\sim} \text{Bernoulli}(\pi), \qquad \pi = \frac{1}{1 + \exp(-\theta)}$$

  - One parameter $\theta$, the log-odds of success.
- The **average** log-likelihood, its score, and its Hessian all have closed forms

  $$\bar{\ell}_n(\theta) = \bar{y}\,\theta - \log\big(1 + e^{\theta}\big), \qquad \bar{S}_n(\theta) = \bar{y} - \pi, \qquad \bar{H}_n(\theta) = -\pi(1-\pi)$$

- The Newton-Raphson step can be written in closed form

  $$\hat{\theta}^{(t+1)} = \hat{\theta}^{(t)} + \frac{\bar{y} - \pi^{(t)}}{\pi^{(t)}\big(1 - \pi^{(t)}\big)}$$

- And the true MLE is also known in closed form: $\hat{\theta} = \log\frac{\bar{y}}{1-\bar{y}}$ 

:::

## Coding the two algorithms

- Simulate some data and write down the three components

```{r, echo=T}
set.seed(60637)
n <- 100
y <- rbinom(n, 1, 1/(1 + exp(-0.8)))   # true theta = 0.8
ybar <- mean(y)

loglik <- function(th) ybar*th - log(1 + exp(th))      # average log-likelihood
score  <- function(th) ybar - 1/(1 + exp(-th))         # average score
hess   <- function(th){ p <- 1/(1 + exp(-th)); -p*(1-p) }  # average Hessian

theta_hat <- log(ybar/(1 - ybar))                      # the analytic MLE
theta_hat
```

## Coding the two algorithms

```{r, echo=T}
newton <- function(th, iter){                 # step = score / curvature
  path <- th
  for (t in 1:iter){ th <- th - score(th)/hess(th); path <- c(path, th) }
  path
}
grad_ascent <- function(th, gamma, iter){     # step = score * fixed learning rate
  path <- th
  for (t in 1:iter){ th <- th + gamma*score(th); path <- c(path, th) }
  path
}
```

## Coding the two algorithms

- Start both at $\hat{\theta}^{(0)} = -1$

```{r, echo=T}
nr_path <- newton(-1, iter = 4)
ga_path <- grad_ascent(-1, gamma = 1, iter = 12)
round(nr_path, 6)
```

```{r, echo=T}
round(ga_path, 6)
```

- Newton-Raphson is at the MLE (`r round(theta_hat,4)`) by the **fourth** iteration.
- Gradient ascent has taken twelve steps and still isn't there.

## Newton-Raphson

```{r, echo=F, fig.width=8.5, fig.height=3.4, fig.align="center"}
grid <- seq(-2.2, 3.2, length.out = 400)
mk <- function(i){
  t0 <- nr_path[i]; t1 <- nr_path[i+1]
  lab <- sprintf("Step %d:  %.2f  to  %.2f", i, t0, t1)
  list(curve = data.frame(panel=lab, th=grid, ll=loglik(grid)),
       quad  = data.frame(panel=lab, th=grid,
                          q = loglik(t0) + score(t0)*(grid-t0) + 0.5*hess(t0)*(grid-t0)^2),
       cur   = data.frame(panel=lab, th=t0, ll=loglik(t0)),
       apex  = data.frame(panel=lab, th=t1,
                          q = loglik(t0) + score(t0)*(t1-t0) + 0.5*hess(t0)*(t1-t0)^2,
                          ll = loglik(t1)))
}
L <- lapply(1:2, mk)
curve_df <- bind_rows(lapply(L,`[[`,"curve")); quad_df <- bind_rows(lapply(L,`[[`,"quad"))
cur_df <- bind_rows(lapply(L,`[[`,"cur"));     apex_df <- bind_rows(lapply(L,`[[`,"apex"))
ggplot() +
  geom_line(data=curve_df, aes(th, ll), linewidth=1) +
  geom_line(data=quad_df, aes(th, q), col="dodgerblue", lty=2, linewidth=.8) +
  geom_point(data=cur_df, aes(th, ll), size=3, col="indianred") +
  geom_point(data=apex_df, aes(th, q), size=3, col="dodgerblue", shape=17) +
  geom_segment(data=apex_df, aes(x=th, xend=th, y=q, yend=ll),
               col="dodgerblue", lty=3, linewidth=.6) +
  geom_point(data=apex_df, aes(th, ll), size=3, col="dodgerblue") +
  geom_vline(xintercept=theta_hat, lty=3, col="grey30") +
  facet_wrap(~panel) + coord_cartesian(ylim=c(-1.65, -0.45)) +
  theme_bw() + xlab(expression(theta)) + ylab("Average log-likelihood")
```

## Gradient ascent

```{r, echo=F, fig.width=8.5, fig.height=3.3, fig.align="center"}
curve_all <- data.frame(th=seq(-2.2,3.2,length.out=400)) %>% mutate(ll=loglik(th))
paths <- bind_rows(
  data.frame(alg="Newton-Raphson",  iter=0:4,  th=nr_path),
  data.frame(alg="Gradient ascent", iter=0:12, th=ga_path)) %>% mutate(ll=loglik(th))
paths$alg <- factor(paths$alg, levels=c("Newton-Raphson","Gradient ascent"))
ggplot() +
  geom_line(data=curve_all, aes(th, ll), linewidth=.9, col="grey40") +
  geom_path(data=paths, aes(th, ll, col=alg), linewidth=.5, alpha=.6) +
  geom_point(data=paths, aes(th, ll, col=alg), size=2.3) +
  geom_vline(xintercept=theta_hat, lty=3) +
  facet_wrap(~alg) +
  scale_color_manual(values=c("Newton-Raphson"="dodgerblue","Gradient ascent"="indianred"), guide="none") +
  coord_cartesian(ylim=c(loglik(-1)-.02, loglik(theta_hat)+.02)) +
  theme_bw() + xlab(expression(theta)) + ylab("Average log-likelihood")
```

## Intro to GLMs

## Motivation: Propensity Scores

::: {.incremental}

- Researchers wanting to estimate causal effects from observational designs often use a *weighting* estimator to account for non-random treatment assignment.
  - Observe treatment $D_i$, confounders $X_i$, outcome $Y_i$
  - Need to estimate the **propensity score** $\pi_i = \Pr(D_i = 1 | X_i)$ to construct "inverse propensity of treatment weights"
- **Key problem**: with many continuous covariates, it's hard to estimate $\Pr(D_i = 1 | X_i)$ non-parametrically
  - One solution: assume a parametric *model* for $D_i$.
- Propensity scores are **predictions** from the model.

:::

## The LaLonde dataset

::: {.incremental}

- @lalonde1986evaluating asked: can selection-on-observables designs recover the answer from a randomized experiment?
  - The **National Supported Work Demonstration (NSW)** randomly assigned disadvantaged workers to a job training program.
  - Outcome: real earnings in 1978 (`re78`).
  - Because assignment was random, the experiment gives a **credible benchmark**: roughly **\$1,794** for the subsample we'll use [@dehejia1999causal].
- LaLonde **created** an observational dataset by replacing the experimental controls with a survey comparison group drawn from the **Panel Study of Income Dynamics** (PSID).
  - Now treatment is *not* randomly assigned -- the comparison group is much older, better educated, and far richer at baseline.
  - If our adjustment methods work, they should recover something near \$1,794 anyway.
- @lalonde1986evaluating found typical OLS regression worked poorly 
  - @dehejia1999causal tackle this with **propensity score** methods (specifically *matching*, but we'll do weighting).

:::

## The LaLonde dataset

- The dataset is in a lot of R packages, such as `causalsens`

```{r, warning=F, message=F, echo=T}
library(causalsens)
data(lalonde.psid)
lalonde <- lalonde.psid   # shorter name for the slides
glimpse(lalonde)
```

- $n = 2{,}675$: **185** NSW treated units and **2,490** PSID comparison units.

## The experimental benchmark

- The **experimental** controls are not in that dataset -- LaLonde replaced them. But they ship alongside it.

```{r, warning=F, message=F, echo=T}
data(lalonde.exp)   # 185 treated + 260 *randomized* controls
lalonde.exp %>% group_by(treat) %>%
  summarize(n = n(), age = mean(age), education = mean(education),
            married = mean(married), re74 = mean(re74))
```

- Randomization did its job -- the two arms look alike on every pre-treatment covariate.

```{r, echo=T}
exp_att <- lm_robust(re78 ~ treat, data = lalonde.exp, se_type = "HC2")
coef(summary(exp_att))["treat",]
```

- Training raised 1978 earnings by about **\$1,794**. That is the number our adjustment methods have to recover.
  - The 185 treated units are the *same people* in both datasets; only the comparison group changes.
  - Note the benchmark is **itself an estimate**, with a 95% CI of roughly $[\$476,\ \$3{,}113]$.

## The naive comparison

- What happens if we just compare treated and control earnings directly?

```{r, echo=T}
lalonde %>% group_by(treat) %>% 
  summarize(n = n(), mean_re78 = mean(re78), mean_re74 = mean(re74), 
            mean_age = mean(age), mean_married = mean(married))
```

- Complete ignorability is implausible -- the PSID men are nine years older, nearly two years better educated, four times as likely to be married, and earned nine times as much in 1974.

## The naive comparison

- Regressing 1978 earnings on treatment alone:

```{r, echo=T}
naive_fit <- lm_robust(re78 ~ treat, data=lalonde, se_type="HC2")
summary(naive_fit)
```

- The training program appears to **lower** earnings by about \$15,205.

## Generalized Linear Models

::: {.incremental}

- Generalized linear models (GLMs) [@nelder1972glm] have three components:
  1) A parametric distribution for $Y_i|X_i$ ("stochastic component")
  2) A linear predictor: $\eta_i = X_i^{\prime}\beta = \beta_0 + \beta_1X_{i1} + \beta_2X_{i2} + \dotsc + \beta_kX_{ik}$ ("systematic component")
  3) A link function $g()$ applied to the CEF $\mathbb{E}[Y_i|X_i]$ that yields the linear predictor

      $$g\big(\mathbb{E}[Y_i | X_i]\big) = \eta_i$$

- Alternatively, we can write the CEF in terms of the "inverse link" $g^{-1}()$ applied to the linear predictor

      $$\mathbb{E}[Y_i | X_i] = g^{-1}(\eta_i)$$

:::

## Exponential Family

::: {.incremental}

- The types of probability distributions permitted for $Y_i$ are quite general: the **exponential family**
  - This contains the Normal as well as many other common distributions -- the Bernoulli, Poisson, exponential, gamma, and more.
- Exponential family distributions have density functions of the form

  $$f(Y|\theta) = h(Y) \exp\bigg\{b(\theta) \cdot T(Y) - A(\theta)\bigg\}$$

- $h(Y)$, $b(\theta)$, $T(Y)$ and $A(\theta)$ are known functions
  - $T(Y)$ is the **sufficient statistic**, $b(\theta)$ the **canonical (natural) parameter**, $A(\theta)$ the **log-normalizer**.
- **The key intuition**: exponential family distributions factorize in a convenient way
  - The parameter and the data interact only through the product $b(\theta) \cdot T(Y)$.
  - If $b(\theta) = \theta$, the distribution is in "canonical" form.

:::

## Example: Bernoulli

::: {.incremental}

- Consider the Bernoulli PMF

  $$f(Y_i|\pi_i) = \pi_i^{Y_i}(1-\pi_i)^{1-Y_i}$$

- Take the log, then the exponent

  $$f(Y_i|\pi_i) = \exp\bigg\{\log\bigg[\pi_i^{Y_i}(1-\pi_i)^{1-Y_i}\bigg]\bigg\}$$

- Properties of logs

  $$f(Y_i|\pi_i) = \exp\bigg\{Y_i\log(\pi_i) + (1-Y_i)\log(1-\pi_i)\bigg\}$$

:::

## Example: Bernoulli

::: {.incremental}

- Rearranging and using properties of logs again

  $$f(Y_i|\pi_i) = \exp\bigg\{Y_i\log\bigg(\frac{\pi_i}{1-\pi_i}\bigg) + \log(1-\pi_i)\bigg\}$$

- So our exponential form is
  - $h(Y) = 1$
  - $T(Y) = Y_i$
  - $b(\pi_i) = \log\bigg(\frac{\pi_i}{1-\pi_i}\bigg)$
  - $A(\pi_i) = -\log(1-\pi_i)$
    - remember the exponential form carries a **minus** sign in front of $A$
- Critically, this is where we get a good link function
  - The **canonical parameter** is $\log\bigg(\frac{\pi}{1-\pi}\bigg)$ -- the log-odds.
  - The **canonical link** is the function that equates this parameter with the linear predictor: $\log\bigg(\frac{\pi_i}{1-\pi_i}\bigg) = X_i^{\prime}\beta$

:::

## Binary outcome models

## Logistic regression

::: {.incremental}

- The "logit" or "logistic" GLM models the **log-odds** of a binary outcome as a function of the linear predictor $X_i^{\prime}\beta$

  $$Y_i \underset{\text{i.i.d}}{\sim} \text{Bernoulli}(\pi_i)$$

  $$\mathbb{E}[Y_i | X_i] = \Pr(Y_i = 1 | X_i) = \pi_i$$

  $$\log\bigg(\frac{\pi_i}{1-\pi_i}\bigg) = X_i^{\prime}\beta$$

- Alternatively, this is written in terms of the "inverse link" function (the logistic function) relating $\pi_i$ to $g^{-1}(X_i^{\prime}\beta)$

  $$\pi_i = \frac{\exp(X_i^{\prime}\beta)}{1 + \exp(X_i^{\prime}\beta)} = \frac{1}{1 + \exp(-X_i^{\prime}\beta)}$$

:::

## Inverse-link functions

- The logistic function maps inputs on $\mathbb{R}$ to $(0, 1)$

```{r, echo=F, fig.width=5, fig.height = 3, fig.align="center"}
logistic <- function(x) 1/(1 + exp(-x))
logistic_curve <- ggplot() + xlim(-5, 5) + ylim(0,1) + geom_function(fun=logistic) + theme_bw() + ylab(expression(pi)) + xlab(expression(paste("X",minute,beta))) + geom_vline(xintercept=0, lty=2) + ggtitle("Logistic function")
logistic_curve
```

## Inverse-link functions

- Another link is the "probit", whose inverse link is the Normal CDF $\Phi()$

```{r, echo=F, fig.width=5, fig.height = 3, fig.align="center"}
normal_cdf <- ggplot() + xlim(-5, 5) + ylim(0,1) + geom_function(fun=pnorm) + theme_bw() + ylab(expression(pi)) + xlab(expression(paste("X",minute,beta))) + geom_vline(xintercept=0, lty=2) + ggtitle("Normal CDF")
normal_cdf
```

## Inverse-link functions

- When do we use probit vs. logit?
  - Computational convenience -- the logit is the canonical link, so it's better behaved.
  - Probit has some good properties for Bayesian inference (we'll come back to this).
- Can you tell the difference?

```{r, echo=F, fig.width=6, fig.height = 3.5, fig.align="center"}
normal_logit <- ggplot() + xlim(-5, 5) + ylim(0,1) + geom_function(fun=pnorm, col="dodgerblue") + geom_function(fun=logistic, lty=2, col="indianred") + theme_bw() + ylab(expression(pi)) + xlab(expression(paste("X",minute,beta))) + geom_vline(xintercept=0, lty=2) + ggtitle("Logistic (red, dashed) vs. Normal CDF (blue)")
normal_logit
```

## Deriving the logit likelihood

::: {.incremental}

- Start from the Bernoulli log-likelihood

  $$\ell_n(\beta) = \sum_{i=1}^n Y_i\log(\pi_i) + (1-Y_i)\log(1 - \pi_i)$$

- Collecting the $Y_i$ terms gives us the log-odds, which is just $X_i^{\prime}\beta$

  $$\ell_n(\beta) = \sum_{i=1}^n Y_i \log\bigg(\frac{\pi_i}{1-\pi_i}\bigg) + \log(1 - \pi_i)$$

- And from the inverse link,

  $$1 - \pi_i = 1 -  \frac{1}{1 + \exp(-X_i^{\prime}\beta)} = \frac{\exp(-X_i^{\prime}\beta)}{1 + \exp(-X_i^{\prime}\beta)} = \frac{1}{1 + \exp(X_i^{\prime}\beta)}$$

- Substituting in

  $$\ell_n(\beta) = \sum_{i=1}^n \bigg[Y_i X_i^{\prime}\beta - \log\bigg(1 + \exp(X_i^{\prime}\beta)\bigg)\bigg]$$

:::

## Example: Estimating the propensity score

- We want to predict assignment to training, `treat`, from the pre-treatment covariates
  - `age` - Age in years
  - `education` - Years of schooling
  - `black`, `hispanic` - Race/ethnicity indicators
  - `married` - Married; binary indicator
  - `nodegree` - No high school degree; binary indicator
  - `re74` - Real earnings in 1974 (pre-treatment)
  - `re75` - Real earnings in 1975 (pre-treatment)

## Example: Estimating the propensity score

- Make the design matrix $\mathbf{X}$

```{r, warning=F, message=F, echo=T}
X_mat <- model.matrix(treat ~ age + education + black + hispanic + married + nodegree + re74 + re75, data=lalonde)
head(X_mat) # View the top of the matrix
```

- How many observations?

```{r, echo=T}
n_obs <- nrow(X_mat)
n_obs
```

## Example: Estimating the propensity score

- Put the log-likelihood into code (this returns a vector of the log-likelihood for each observation)

```{r, warning = F, message = F}
logit_loglik <- function(beta, Y, X){
  eta <- X%*%beta # linear predictor
  lik <- Y*eta - log(1+exp(eta))
  return(lik)
}
```

- We derived the score analytically, so let's supply it too -- one row per observation

```{r, warning = F, message = F}
logit_grad <- function(beta, Y, X){
  eta <- X%*%beta
  pi_hat <- 1/(1 + exp(-eta))
  X * as.vector(Y - pi_hat)  
}
```

## Example: Estimating the propensity score

- Now let's optimize it to get the MLE

```{r, warning=F, message=F}
library(maxLik)

logit_mle <- maxLik(logit_loglik, 
                      grad = logit_grad,
                      Y=lalonde$treat,
                      X=X_mat,
                      start = rep(0, ncol(X_mat)),
                      method = "NR")
```

- What did we get?

```{r, echo=T}
est <- coef(logit_mle) # Our optimization routine
names(est) <- colnames(X_mat)
est
```

## Example: Estimating the propensity score

- Compare to the built-in R routine

```{r, echo=T}
logit_Rglm <- glm(treat ~ age + education + black + hispanic + married + nodegree + re74 + re75, 
                  data=lalonde, family=binomial(link="logit"))
coef(logit_Rglm)
```

- Let's obtain our (asymptotic) variance-covariance matrix -- the inverse of the observed information

```{r, echo=T}
logit_vcov <- solve(-maxLik::hessian(logit_mle))
```

- Square root of the diagonal gives our SEs

```{r, echo=T}
logit_SEs <- sqrt(diag(logit_vcov))
```

## Example: Estimating the propensity score

- Let's get our test statistics and p-values, using the usual asymptotic normal approximation

```{r, echo=T}
results <- rbind(coef(logit_mle), logit_SEs, 
                     coef(logit_mle)/logit_SEs, 
                     2*pnorm(-abs(coef(logit_mle)/logit_SEs)))
colnames(results) <- colnames(X_mat)
rownames(results) <- c("Estimate", "Std. Error", "Test statistic", "p-value")
round(results, 4)
```

## Interpreting logit coefficients

::: {.incremental}

- How do we interpret the $\beta$ parameters substantively?

  $$\log\bigg(\frac{\pi_i}{1-\pi_i}\bigg) = \beta_0 + \beta_1X_{i1} + \beta_2X_{i2} + \dotsc + \beta_k X_{ik}$$

- Take the partial derivative w.r.t. $X_{ik}$

  $$\frac{\partial}{\partial X_{ik}} \log\bigg(\frac{\pi_i}{1-\pi_i}\bigg) =  \beta_k$$

- So $\beta_k$ captures the change in the log-odds for a one-unit change in $X_{k}$
  - Descriptively, it's the difference in log-odds between two observations that differ in $X_{k}$ by one unit.
  - Exponentiating, $\exp(\beta_k)$ is an **odds ratio**. This is *not* a risk ratio, and it is not a difference in probabilities.

:::

## Interpreting logit coefficients

::: {.incremental}

- On the "log-odds" scale, the change due to an increase in $X_{k}$ does not depend on the values of the other $X$ variables (unless we explicitly specify an interaction).
  - But thinking on the log-odds scale is hard! We think in terms of probabilities.
  - This additivity *does not hold* when we take $\frac{\partial}{\partial X_{i1}} \pi_i$
- Working through the chain rule,

  $$\frac{\partial \pi_i}{\partial X_{ik}} = \beta_k \, \pi_i(1 - \pi_i)$$

  - The effect on the probability scale depends on $\pi_i$, and therefore on **every** covariate.
- Logit models *implicitly* encode interactions with respect to the CEF $\mathbb{E}[Y_i | X_i]$.

:::

## Interpreting logit coefficients

- **Remember**: A one-unit change in the linear predictor corresponds to *different* changes in probability depending on your baseline.

```{r, echo=F, fig.width=5, fig.height = 3, fig.align="center"}
#| layout-ncol: 2

logistic <- function(x) 1/(1 + exp(-x))
draw_points <- data.frame(x1 = c(0, 1), x2 = c(0, 1), y1 = c(0,0), y2 = logistic(c(0,1)))
draw_points2 <- data.frame(x1 = c(-5, -5), x2 = c(0, 1), y1 = logistic(c(0,1)), y2 = logistic(c(0,1)))
logistic_marginal <- ggplot() + xlim(-5, 5) + ylim(0,1) + geom_function(fun=logistic) + theme_bw() + ylab(expression(pi)) + 
  xlab(expression(paste("X",minute,beta))) + geom_vline(xintercept=0, lty=2) + ggtitle("Marginal Effect at 0") +
  geom_segment(aes(x=x1, y=y1, xend=x2, yend=y2), data=draw_points, col="dodgerblue", lwd=2) +
  geom_segment(aes(x=x1, y=y1, xend=x2, yend=y2), data=draw_points2, col="indianred", lwd=2) 
  
logistic_marginal

draw_points_2 <- data.frame(x1 = c(2, 3), x2 = c(2, 3), y1 = c(0,0), y2 = logistic(c(2,3)))
draw_points_22 <- data.frame(x1 = c(-5, -5), x2 = c(2, 3), y1 = logistic(c(2,3)), y2 = logistic(c(2,3)))
logistic_marginal_2 <- ggplot() + xlim(-5, 5) + ylim(0,1) + geom_function(fun=logistic) + theme_bw() + ylab(expression(pi)) + 
  xlab(expression(paste("X",minute,beta))) + geom_vline(xintercept=0, lty=2) + ggtitle("Marginal Effect at 2") +
  geom_segment(aes(x=x1, y=y1, xend=x2, yend=y2), data=draw_points_2, col="dodgerblue", lwd=2) +
  geom_segment(aes(x=x1, y=y1, xend=x2, yend=y2), data=draw_points_22, col="indianred", lwd=2) 
  
logistic_marginal_2
```

## Inference on transformations

## Transformed quantities

::: {.incremental}

- We have $\hat{\beta}$, but we actually want the **propensity score** $\hat{\pi}_i = \Pr(D_i = 1 | X_i)$
  - Just apply the inverse link to get the quantity we want

  $$\hat{\pi}_i = \frac{1}{1 + \exp(-X_i^{\prime}\hat{\beta})}$$

- By **invariance**, this is the MLE of $\pi_i$, and it's consistent for the true propensity score (under our modeling assumptions).
- But if we want a confidence interval for $\hat{\pi}_i$, we use the **delta method**.

:::

## Example: a predicted propensity

- Let's get the propensity of receiving training for a **typical NSW participant** -- the median covariate profile among the treated

```{r}
X_medians <- apply(X_mat[lalonde$treat == 1, ], 2, median)
X_medians
```

- Construct our prediction function

```{r}
pred_prob <- function(beta, X){
  return(1/(1 + exp(-X%*%beta)))
}
pred_median <- pred_prob(coef(logit_mle), X_medians)
pred_median
```

## Example: a predicted propensity

- Let's do the gradient numerically (we could calculate it analytically though!)

```{r}
pred_prob_gradient <- numericGradient(pred_prob, 
                                      t0 = coef(logit_mle),
                                      X = X_medians)
```

- Applying the delta method

```{r}
pred_prob_var <- pred_prob_gradient%*%logit_vcov%*%t(pred_prob_gradient) # R treats vectors as row-vectors not column vectors
```

- Making our 95% asymptotic CI for $\hat{\pi}(X)$

```{r}
c(pred_median - abs(qnorm(.025))*sqrt(pred_prob_var),
  pred_median + abs(qnorm(.025))*sqrt(pred_prob_var))
```

## `marginaleffects`

- The `marginaleffects` R package provides a straightforward implementation of the delta method for a lot of transformed quantities
  - (e.g.) the marginal change in the **probability of** $D_i=1$ for a one-unit change in $X$
- `avg_predictions()` will calculate point estimates + CIs for the expected outcome just like we did above.
- `avg_comparisons()` will do the implied "marginal effects" for a unit change in $X$ given a set baseline.
- We have to tell it **which covariate profile** to evaluate at. We'll use the **treated** units -- the NSW participants are the group we care about.

```{r, echo=T, message=F}
library(marginaleffects)
treated <- subset(lalonde, treat == 1)

grid_median <- datagrid(model = logit_Rglm, newdata = treated,
                        FUN_numeric = median, FUN_integer = median, FUN_binary = median)
grid_mean   <- datagrid(model = logit_Rglm, newdata = treated,
                        FUN_numeric = mean,   FUN_integer = mean,   FUN_binary = mean)
```

## `marginaleffects`

- The **median** profile is the one we just built by hand; the **mean** profile is the other common convention

```{r, echo=T}
rbind(median = unlist(grid_median[, -1]), mean = unlist(grid_mean[, -1]))
```

::: {.incremental}

- Note the mean profile is not a person anyone could be -- it is 84% Black and 71% without a degree.
  - The median profile at least corresponds to a *possible* respondent.

:::

## `marginaleffects`: predicted propensity

```{r, echo=T}
avg_predictions(logit_Rglm, newdata = grid_median)   # matches our hand calculation
```

```{r, echo=T}
avg_predictions(logit_Rglm, newdata = grid_mean)
```

- Evaluating at the *mean* moves us from $\hat{\pi} \approx 0.89$ down to $\approx 0.64$ -- the profile is different, so the point on the curve is different.

## `marginaleffects`: marginal effect of `married`

```{r, echo=T}
avg_comparisons(logit_Rglm, variables = "married", newdata = grid_median)
```

```{r, echo=T}
avg_comparisons(logit_Rglm, variables = "married", newdata = grid_mean)
```

::: {.incremental}

- Being married lowers the probability of NSW participation by **28** points at the median profile, **39** at the mean profile.
  - Same $\hat{\beta}$, same data -- only the baseline $\pi_i$ differs. This is the $\beta_k\pi_i(1-\pi_i)$ scaling at work.

:::


## Inverse propensity of treatment weighting

::: {.incremental}

- Now we have $\hat{\pi}_i$ for every unit.
- The control group is unrepresentative of the treated group.
  - So we **reweight** it. Upweight controls who look like treated units; downweight the ones who don't.
- Under **conditional ignorability** ($\{Y_i(1), Y_i(0)\} \perp D_i \mid X_i$) and **positivity** ($0 < \pi_i < 1$), the IPTW estimator identifies a causal contrast [@rosenbaum1983central].
- For the **average treatment effect on the treated** (ATT), the weights are

  $$w_i = D_i + (1 - D_i)\frac{\hat{\pi}_i}{1 - \hat{\pi}_i}$$

  - Treated units get weight 1 -- we keep the treated sample as-is and re-weight controls to match.
  - Controls get the **odds** of treatment: a control who looked very likely to be treated counts for more.

:::

## Inverse propensity of treatment weighting

::: {.incremental}

- The ATT estimator is then a difference of weighted means

  $$\widehat{\text{ATT}} = \frac{1}{n_1}\sum_{i: D_i = 1} Y_i \ - \ \frac{\sum_{i: D_i = 0} w_i Y_i}{\sum_{i: D_i = 0} w_i}$$

- Dividing by the **sum of weights** rather than $n_0$ gives the *Hájek* (stabilized) estimator
- For the **ATE** instead, we would weight *every* unit by the inverse probability of the treatment it actually received

  $$w_i^{\text{ATE}} = \frac{D_i}{\hat{\pi}_i} + \frac{1 - D_i}{1 - \hat{\pi}_i}$$

- We'll target the ATT, since that's what the experimental benchmark of \$1,794 estimates.

:::

## Checking overlap first

- Before weighting anything, look at the estimated propensity scores by treatment status

```{r, echo=T}
lalonde$pscore <- as.vector(pred_prob(coef(logit_mle), X_mat))
```

```{r, echo=F, fig.width=7.5, fig.height=3.6, fig.align="center"}
lalonde %>% mutate(grp = factor(ifelse(treat==1, "Treated (NSW), n = 185", "Control (PSID), n = 2,490"),
                                levels=c("Treated (NSW), n = 185", "Control (PSID), n = 2,490"))) %>%
  ggplot(aes(x=pscore, fill=grp)) +
  geom_histogram(bins=50) +
  facet_wrap(~grp, ncol=1, scales="free_y") +   # free y: the control spike would hide the treated
  scale_fill_manual(values=c("Treated (NSW), n = 185"="dodgerblue",
                             "Control (PSID), n = 2,490"="indianred"), guide="none") +
  theme_bw() + xlab("Estimated propensity score") + ylab("Count")
```

- A large mass of PSID controls has $\hat\pi \approx 0$, not close to the treated group.

## Checking overlap first

::: {.incremental}

- Effectively, weighting **discards** most of the PSID sample.

```{r, echo=T}
lalonde$w_att <- with(lalonde, ifelse(treat == 1, 1, pscore/(1 - pscore)))
c(max_control_weight = max(lalonde$w_att[lalonde$treat==0]),
  sum_control_weights = sum(lalonde$w_att[lalonde$treat==0]))
```

:::

## Did the weights fix the imbalance?

- The point of the weights is **balance**. After weighting, the covariate distributions should look alike.
- The standard diagnostic is the **standardized mean difference**

  $$\text{SMD}_k = \frac{\bar{X}_{k,\text{treated}} - \bar{X}_{k,\text{control}}^{\,w}}{s_{k,\text{treated}}}$$

  - Scaling by the treated-group SD puts every covariate on a common, unit-free scale.
  - Common heuristic for observational studies (from biostats/epi): $|\text{SMD}| < 0.1$ is "balanced."
- The `cobalt` package does this.

## Did the weights fix the imbalance?

```{r, echo=T, message=F, warning=F}
library(cobalt)
bt <- bal.tab(treat ~ age + education + black + hispanic + married + nodegree + re74 + re75,
              data = lalonde, weights = lalonde$w_att,
              method = "weighting", estimand = "ATT",
              un = TRUE, binary = "std", thresholds = c(m = .1))
bt$Balance
```

## Did the weights fix the imbalance?

- And the **effective sample sizes**:

```{r, echo=T}
bt$Observations
```

::: {.incremental}

- The unweighted imbalances are enormous -- `re74` starts at $-3.5$ standard deviations, `married` at $-1.7$.
  - After weighting nearly everything is inside or roughly at the $0.1$ band.
- **Note the effective sample size**: the 2,490 controls are worth about **43** after weighting.
  - This is what our histogram shows - weighting puts a large amount of weight on a small number of units.
  - Low effective sample size = higher variance estimator.
- Don't forget that we by definition can't check balance on things we **didn't** control for.
  - Balance on the covariates we *included* says nothing about the confounders we omitted.

:::

## The Love plot

- Common to present this in a **Love plot**

```{r, echo=T, message=F, warning=F, fig.width=8, fig.height=3.6, fig.align="center"}
love.plot(treat ~ age + education + black + hispanic + married + nodegree + re74 + re75,
          data = lalonde, weights = lalonde$w_att,
          method = "weighting", estimand = "ATT",
          binary = "std", thresholds = c(m = .1),
          var.names = c(age = "Age", education = "Years of education", married = "Married",
                        nodegree = "No HS degree", re74 = "Earnings 1974",
                        re75 = "Earnings 1975", black = "Black", hispanic = "Hispanic"),
          sample.names = c(Unadjusted = "Unweighted", Adjusted = "IPTW (ATT)"),
          colors = c("indianred", "dodgerblue"), shapes = c("circle", "triangle"),
          title = "Covariate balance before and after weighting")
```

- Red circles are the unweighted sample, blue triangles the weighted one. Dashed lines mark $\pm 0.1$.

## The IPTW estimate

- Now compute the weighted difference in means

```{r, echo=T}
att_hajek <- with(lalonde,
  mean(re78[treat == 1]) - weighted.mean(re78[treat == 0], w_att[treat == 0]))
att_hajek
```

## The IPTW estimate

- Equivalently, as a weighted least squares regression

```{r, echo=T}
summary(lm_robust(re78 ~ treat, data = lalonde, weights = w_att, se_type = "HC2"))
```

## The IPTW estimate

- Compare to the alternatives:

```{r, echo=T}
naive_est <- with(lalonde, mean(re78[treat==1]) - mean(re78[treat==0]))
reg_fit <- lm_robust(re78 ~ treat + age + education + black + hispanic + married + nodegree + re74 + re75,
                     data=lalonde, se_type="HC2")
reg_est <- coef(reg_fit)["treat"]
data.frame(estimator = c("Naive difference", "Regression adjustment", "IPTW (ATT)", "Experimental benchmark"),
           estimate  = c(naive_est, as.numeric(reg_est), att_hajek, coef(exp_att)["treat"]))
```

## Estimating the standard errors

::: {.incremental}

- For the standard errors, we want to account for estimation uncertainty in the propensity scores.
- Two options:
  - **Bootstrap** the whole procedure -- refit the logit *inside* each bootstrap iteration.
  - Stack the score functions of both stages and apply the **sandwich estimator**
- We'll implement the bootstrap

```{r, echo=T}
att_fn <- function(dd){
  Xb <- model.matrix(treat ~ age + education + black + hispanic + married + nodegree + re74 + re75, data=dd)
  fit <- suppressWarnings(glm.fit(Xb, dd$treat, family = binomial()))
  ps <- as.vector(1/(1 + exp(-Xb %*% fit$coefficients)))
  w  <- ifelse(dd$treat == 1, 1, ps/(1 - ps))
  mean(dd$re78[dd$treat == 1]) - weighted.mean(dd$re78[dd$treat == 0], w[dd$treat == 0])
}
set.seed(53703)
boot_att <- replicate(2000, att_fn(lalonde[sample(nrow(lalonde), replace = TRUE),]))
c(bootstrap_SE = sd(boot_att), quantile(boot_att, c(.025, .975)))
```

:::

## Results

- Collect the four estimators, each with its own standard error

```{r, echo=T}
grab <- function(f) coef(summary(f))["treat", c("Estimate", "Std. Error", "CI Lower", "CI Upper")]
results_tab <- data.frame(
  estimator = c("Naive difference", "Regression adjustment", "IPTW (ATT)", "Experimental benchmark"),
  rbind(grab(naive_fit), grab(reg_fit),
        c(att_hajek, sd(boot_att), unname(quantile(boot_att, c(.025, .975)))),
        grab(exp_att)),
  se = c("HC2", "HC2", "bootstrap", "HC2"),
  row.names = NULL)
names(results_tab) <- c("estimator", "estimate", "std_error", "ci_lower", "ci_upper", "se_type")
```

## Results

```{r, echo=T}
results_tab
```

::: {.incremental}

- Adjustment moved us an enormous distance: $-\$15{,}205 \rightarrow +\$1{,}759$, against a benchmark of $\$1{,}794$.
  - Regression adjustment alone only reaches about $+\$752$
- But the bootstrap interval is **wide** and includes zero.
  - With the comparison group weighted down to an effective 43 units, there simply isn't much information left.
- IPTW estimators are **extremely noisy** when propensities are close to $0$ or $1$.
  - **Doubly-robust** approaches (e.g. AIPW) combine the weights with an outcome model, which helps reduce variance [@kang2007demystifying].

:::


## "Latent variables" and logit

## Latent variables

::: {.incremental}

- Another common way of formulating discrete outcome regressions is in terms of an unobserved continuous "latent" variable and an observation mechanism.
  - Instead of putting a distribution on $Y_i$, we put one on an unobserved $Y_i^*$
  - $Y_i$ is some function of $Y_i^*$ (usually an indicator function of some sort)
- For the logistic regression

  $$Y_i = \mathbb{1}(Y_i^* \ge 0)$$
  $$Y_i^* = X_i^{\prime}\beta + \epsilon_i$$
  $$\epsilon_i \sim \text{Logistic}(0, 1)$$

- The Logistic distribution is parameterized in terms of a "location" (mean) $\mu$ and a "scale" parameter $s>0$.
  - For $\mu = 0$, $s= 1$ we have the "standard" logistic

  $$\Pr(\epsilon_i \le x) = \frac{1}{1 + \exp(-x)}$$

:::

## Latent variables

::: {.incremental}

- So what's $\mathbb{E}[Y_i|X_i] = \Pr(Y_i = 1|X_i)$?

  $$\Pr(Y_i = 1|X_i) = \Pr(Y_i^* \ge 0 | X_i) = \Pr(\epsilon_i \ge -X_i^{\prime}\beta) = \Pr(\epsilon_i \le X_i^{\prime}\beta) = \frac{1}{1 + \exp(-X_i^{\prime}\beta)}$$

  - The last-but-one step uses the **symmetry** of the logistic distribution around zero.
- We can write the **probit** similarly, defining the error distribution as

  $$\epsilon_i \sim \text{Normal}(0, 1)$$

- And by extension, the probability $\Pr(Y_i = 1|X_i)$

  $$\Pr(Y_i = 1|X_i) = \Pr(Y_i^* \ge 0|X_i) = \Pr(\epsilon_i \le X_i^{\prime}\beta) = \Phi(X_i^{\prime}\beta)$$

- **Note** the scale of $\epsilon_i$ is fixed by assumption, not estimated
  - $Y_i^*$ is never observed, so only the *ratio* $\beta/s$ is identified -- so we normalize to $s =1$

:::

## Latent variables

- We can visualize this as an integral over the density of the random "error" $\epsilon_i$

```{r, echo=F, fig.width=5, fig.height = 3, fig.align="center"}
logistic_dens <- ggplot() + xlim(-5, 5) + geom_function(fun=dlogis) + theme_bw() + ylab("Density") + geom_area(aes(x=c(-5,2)),stat = "function", fun = dlogis, fill = "dodgerblue", xlim = c(-5, 2)) + 
  xlab(expression(epsilon)) + geom_vline(xintercept=0, lty=2) + geom_vline(xintercept=2, lty=3, col="indianred", lwd=1) 

logistic_dens
```

## Latent variables

- Same thing for a probit

```{r, echo=F, fig.width=5, fig.height = 3, fig.align="center"}
probit_dens <- ggplot() + xlim(-5, 5) + geom_function(fun=dnorm) + theme_bw() + ylab("Density") + geom_area(aes(x=c(-5,2)),stat = "function", fun = dnorm, fill = "dodgerblue", xlim = c(-5, 2)) + 
  xlab(expression(epsilon)) + geom_vline(xintercept=0, lty=2) + geom_vline(xintercept=2, lty=3, col="indianred", lwd=1) 

probit_dens
```

## Ordinal logit

::: {.incremental}

- Using the latent variable formulation lets us extend binary outcome models to outcomes with multiple discrete levels.
- For example, consider an "ordinal" response variable
  - (e.g.) a Likert scale (1 = "strongly disagree", 3 = "neutral", 5 = "strongly agree")
- The responses are *ordered* but the intervals between them do not have any meaning.
  - A popular modeling technique with an ordinal $Y$ is to treat it as an expression of a latent variable.

:::

## Ordinal logit

::: {.incremental}

- In an *ordered logit* model, an ordinal outcome $Y$ with $L$ unique, ordered values $Y \in \{1, 2, 3, \dotsc L\}$ is a function of a latent variable and a set of $L-1$ cutpoints $\kappa$

  $$Y_i = \begin{cases}1 &\text{if } Y_i^* \le \kappa_1 \\
                        2 &\text{if } \kappa_1 < Y_i^* \le \kappa_2\\
                        3 &\text{if } \kappa_2 < Y_i^* \le \kappa_3\\
                        \vdots &\\
                        L &\text{if } Y_i^* > \kappa_{L-1}\end{cases}$$

- We still assume $Y_i^*$ has a logistic distribution

  $$Y_i^* = X_i^{\prime}\beta + \epsilon_i, \qquad \epsilon_i \sim \text{Logistic}(0, 1)$$

:::

## Ordinal logit

::: {.incremental}

- To estimate the model, we now have to estimate both the $\beta$ parameters and the $\kappa$ parameters
  - Note we omit the intercept from $X_i$ -- the $\kappa$ parameters are essentially the "intercepts" for each boundary between the choices.
- How does the model map changes in the latent variable to changes in probability? Work with the **cumulative** probabilities:

  $$\Pr(Y_i \le l \,|\, X_i) = \Pr(Y_i^* \le \kappa_l \,|\, X_i) = \Pr(\epsilon_i \le \kappa_l - X_i^{\prime}\beta)$$

- And then difference adjacent cumulatives to get each cell probability

  $$\Pr(Y_i = l \,|\, X_i) = \Pr(Y_i \le l \,|\, X_i) - \Pr(Y_i \le l-1 \,|\, X_i)$$

  with the conventions $\Pr(Y_i \le 0) = 0$ and $\Pr(Y_i \le L) = 1$.

:::

## Ordinal logit

- Let's think in terms of the distribution of $Y_i^*$ and the cut-points for a case with $L=3$. For $X_i^{\prime}\beta = 0$:

```{r, echo=F, fig.width=7, fig.height = 3, fig.align="center"}
logistic_dens <- ggplot() + xlim(-8, 8) + geom_function(fun=dlogis) + theme_bw() + ylab("Density") + geom_area(aes(x=c(-8,-1)),stat = "function", fun = dlogis, fill = "dodgerblue", xlim = c(-8, -1)) +  geom_area(aes(x=c(-1,2)),stat = "function", fun = dlogis, fill = "darkorange", xlim = c(-1, 2)) + geom_area(aes(x=c(2,8)),stat = "function", fun = dlogis, fill = "gold", xlim = c(2, 8)) + 
  xlab(expression(Y^"*")) + geom_vline(xintercept=0, lty=2) + geom_vline(xintercept=-1, lty=3, col="darkgreen", lwd=1) + geom_vline(xintercept=2, lty=3, col="darkgreen", lwd=1) 

logistic_dens
```

- The green lines are the cutpoints $\kappa_1, \kappa_2$; the shaded areas are $\Pr(Y_i = 1), \Pr(Y_i = 2), \Pr(Y_i = 3)$.

## Ordinal logit

- Shifting the mean of $Y^*$ up by 3:

```{r, echo=F, fig.width=7, fig.height = 3, fig.align="center"}
logistic_dens <- ggplot() + xlim(-8, 8) + geom_function(fun=dlogis, args=list(location=3)) + theme_bw() + ylab("Density") + geom_area(aes(x=c(-8,-1)),stat = "function", fun = dlogis, fill = "dodgerblue", xlim = c(-8, -1),  args=list(location=3)) +  geom_area(aes(x=c(-1,2)),stat = "function", fun = dlogis, fill = "darkorange", xlim = c(-1, 2),  args=list(location=3)) + geom_area(aes(x=c(2,8)),stat = "function", fun = dlogis, fill = "gold", xlim = c(2, 8),  args=list(location=3)) + 
  xlab(expression(Y^"*")) + geom_vline(xintercept=3, lty=2) + geom_vline(xintercept=-1, lty=3, col="darkgreen", lwd=1) + geom_vline(xintercept=2, lty=3, col="darkgreen", lwd=1) 

logistic_dens
```

- The cutpoints stay put but the *distribution* slides. All three probabilities move together as a result.

## Ordinal logit

::: {.incremental}

- This structural mapping from $Y^*$ to $\Pr(Y_i = l)$ encodes a particular assumption about how these probabilities relate to one another.
- Referred to as the **proportional odds** assumption

  $$\log\left(\frac{\Pr(Y_i \le l)}{1 - \Pr(Y_i \le l)}\right) = \kappa_l - X_i^{\prime}\beta$$

  - The same $\beta$ appears for *every* $l$; only the intercept $\kappa_l$ changes.
- **Intuitively**: the $\beta$ parameters are shared across all the possible outcome levels. Information about whether someone responds "strongly agree" *also* tells us something about how likely they would be to respond "agree", "neutral", "disagree", or "strongly disagree".
  - Not innocuous if there's non-monotonicity or heterogeneous effects.
  - Be careful with these models!

:::

## Multinomial logit

::: {.incremental}

- In a setting with multiple **unordered** outcomes, the natural extension of the logit regression is the **multinomial** logit
- Suppose we have an outcome with $K$ unordered levels (e.g. "choices"). Define a latent "utility" $Y_{ik}^*$ for each of the $K$ categories
  - For identification, we set one category to be the "baseline" by anchoring its coefficients to zero: $\beta_1 = 0$

  $$\begin{align*}Y_{i1}^* &= X_i^{\prime}\beta_1 + \epsilon_{i1} = \epsilon_{i1}\\
  Y_{i2}^* &= X_i^{\prime}\beta_2 + \epsilon_{i2}\\
  &\vdots \\
  Y_{iK}^* &= X_i^{\prime}\beta_K + \epsilon_{iK}\end{align*}$$

- The probability of observing choice $Y_{i} = k$ is the probability that the $k$th latent variable is greater than all of the others

  $$\Pr(Y_i = k \,|\, X_i) = \Pr\big(Y^*_{ik} = \max\{Y^*_{i1}, Y^*_{i2}, \dotsc Y^*_{iK}\}\big)$$

:::

## Multinomial logit

::: {.incremental}

- It turns out that if the error terms $\epsilon_{ik}$ have i.i.d. "Type I extreme value" distributions, we can write the choice probabilities in closed form

  $$\Pr(Y_i = k \,|\, X_i) = \frac{\exp(X_i^{\prime}\beta_{k})}{\sum_{l=1}^K \exp(X_i^{\prime}\beta_{l})}$$

- This is sometimes referred to as the "softmax" function in machine learning.
- **Note**: with the normalized baseline category $\beta_1 = 0$, we can write this equivalently as

  $$\Pr(Y_i = k \,|\, X_i) = \frac{\exp(X_i^{\prime}\beta_{k})}{1 + \sum_{l=2}^K \exp(X_i^{\prime}\beta_{l})}$$

- The i.i.d. error assumption implies **independence of irrelevant alternatives** (IIA).
  - The relative odds of choice $k$ vs. $l$ don't depend on what *other* options are available.
  - Famously implausible in some choice settings (the "red bus/blue bus" problem).

:::

## Example: Appointer Affiliation Bias

- @puig2017affiliation ran a survey experiment on a group of international arbitration experts to see how manipulating the appointing party affected their decisionmaking in a vignette
  - Respondents were given a hypothetical dispute and asked to determine who should pay the legal costs
  - Three options: **Loser pays all**, **Loser pays some**, **Both sides split**

## Example: Appointer Affiliation Bias

- Read in the responses -- `-99` is the survey platform's missing-value code

```{r, echo=T, message = F}
study1 <- read_csv("data/experiment1_cleaned.csv", na = c("", "NA", "-99"))
```

- The outcome is stored as an integer, so recode it into an **ordered factor**
  - The ordering is what makes this an ordinal outcome rather than a nominal one

```{r, echo=T}
study1 <- study1 %>%
  mutate(isReimbursementFolded_Factor = case_when(
    isReimbursementFolded == 1 ~ "Winner pays all",
    isReimbursementFolded == 2 ~ "Winner pays some",
    isReimbursementFolded == 3 ~ "Split costs",
    isReimbursementFolded == 4 ~ "Loser pays some",
    isReimbursementFolded == 5 ~ "Loser pays all"))

study1$isReimbursementFolded_Factor <- factor(
  study1$isReimbursementFolded_Factor,
  levels = c("Winner pays all", "Winner pays some", "Split costs",
             "Loser pays some", "Loser pays all"), ordered = TRUE)
```

## Example: Appointer Affiliation Bias

- Keep the completed responses, and drop the two "winner pays" categories
  - Essentially no one chose them, so we're left with the **three** options in the vignette

```{r, echo=T}
study1_complete <- study1 %>%
  filter(!is.na(isReimbursementFolded) & Finished == 1) %>%
  filter(isReimbursementFolded_Factor != "Winner pays some" &
         isReimbursementFolded_Factor != "Winner pays all")

table(study1_complete$isReimbursementFolded_Factor)
```

## Example: Appointer Affiliation Bias

```{r, echo=F, fig.align="center", fig.width=7, fig.height=3.2}
study1_complete %>% ggplot(aes(x=isReimbursementFolded_Factor)) + geom_bar() + xlab("Selected outcome") + ylab("Count") + ggtitle("Distribution of outcome in Puig and Strezhnev (2017)") + theme_minimal()
```

## Example: Appointer Affiliation Bias

- It would seem like an ordinal logit regression would be appropriate here -- a three-category discrete outcome with a clear ordering
  - Implemented in the `ordinal` package

```{r, echo=T}
ordered_logit <- ordinal::clm(isReimbursementFolded_Factor ~ isAppointerSide, data=study1_complete %>% 
                       filter(isAppointerSide == "Appointed by Winner"|isAppointerSide == "Appointed by Loser"))
```

```{r, echo=T}
summary(ordered_logit)
```

## Example: Appointer Affiliation Bias

- The model implies that arbitrators appointed by the winning party shift mostly from **splitting the costs** to making the loser pay for **everything**
  - But is this real or an artifact of the proportional odds assumption? Is there a "constant effect" on the latent variable scale?

```{r}
marginaleffects::avg_comparisons(ordered_logit)
```

## Example: Appointer Affiliation Bias

- Compare with **multinomial logit** (which in this case is equivalent to OLS for each category since we're using a fully-saturated regression)

```{r, echo = T, warning = F, message = F, results = "hide"}
multinom_logit <- nnet::multinom(isReimbursementFolded_Factor ~ isAppointerSide, data=study1_complete %>% 
                       filter(isAppointerSide == "Appointed by Winner"|isAppointerSide == "Appointed by Loser"))
```

```{r, echo=T}
marginaleffects::avg_comparisons(multinom_logit)
```

- Running the ordinal logit (as a reviewer asked us to do!) would have given us the **wrong result**
  - Remember, don't just fixate on the outcome when choosing a modeling strategy - think about what your entire bundle of assumptions implies!


## Next week

::: {.incremental}

- **More likelihood models**
- **Duration models**
  - The survival function, the hazard, and censoring
  - Parametric (exponential, Weibull), semi-parametric (Cox), and non-parametric (Kaplan-Meier) approaches
- **Event count models**
  - Poisson regression, overdispersion, and "agnostic" Poisson (quasi-MLE)
- **Readings:** Chapter 15 of @gelman2020ros; @clark2003survival; @stensrud2020proportional; @wooldridge1999quasi

:::

## References {.refs-slide}

::: {#refs}
:::
