---
title: "Week 4: More Likelihood Models"
subtitle: "PS 818 - Data Analysis with Statistical Models"
author:
  - name: "Anton Strezhnev"
    affiliation: "University of Wisconsin-Madison"
    email: strezhnev@wisc.edu
date: "September 28, 2026"
date-format: "MMMM D, YYYY"
editor: 
  render-on-save: true
format: 
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    overview: true
    navigation-mode: vertical
    progress: false
    controls: true
    link-external-newwindow: true
    footer: "PS 818 - University of Wisconsin-Madison"
    html-math-method:
      method: mathjax
      url: https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js
execute: 
  cache: false
embed-resources: true
bibliography: ../ps818.bib
link-citations: true

---

## More Likelihood Models {.title-slide}

::: {style="display: none"}
$$
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}
$$
:::

```{r, echo = F, message=F, warning=F}
library(tidyverse)
library(knitr)
library(broom)
options(digits=3)
```

## Previously

::: {.incremental}

- **Generalized linear models**

  $$g(\mathbb{E}[Y_i|X_i]) = X_i^{\prime}\beta$$

- **Binary outcome models**
  - Careful with interpreting the raw model coefficients
  - For a logit, $\beta_k$ is a change in **log-odds**, and $\frac{\partial \pi_i}{\partial X_{ik}} = \beta_k\pi_i(1-\pi_i)$
- **Inference on transformations**:
  - Invariance + **delta method**

:::

## This week

::: {.incremental}

- **Latent variables and discrete choice**
  - The latent-variable representation of the logit and probit
  - **Ordinal** logit and the proportional odds assumption
  - **Multinomial** logit and independence of irrelevant alternatives (IIA)
- **Random utility and conditional logit**
  - Choice data and alternative-specific covariates
  - The **model based** approach to **conjoint experiments**
- **Duration models**
  - Survival functions, the hazard, and censoring
  - Parametric, semi-parametric (Cox), and non-parametric (Kaplan-Meier) approaches

:::

## "Latent variables" and logit

## Latent variables

::: {.incremental}

- Another common way of formulating discrete outcome regressions is in terms of an unobserved continuous "latent" variable and an observation mechanism.
  - Instead of putting a distribution on $Y_i$, we put one on an unobserved $Y_i^*$
  - $Y_i$ is some function of $Y_i^*$ (usually an indicator function of some sort)
- For the logistic regression

  $$Y_i = \mathbb{1}(Y_i^* \ge 0)$$
  $$Y_i^* = X_i^{\prime}\beta + \epsilon_i$$
  $$\epsilon_i \sim \text{Logistic}(0, 1)$$

- The Logistic distribution is parameterized in terms of a "location" (mean) $\mu$ and a "scale" parameter $s>0$.
  - For $\mu = 0$, $s= 1$ we have the "standard" logistic

  $$\Pr(\epsilon_i \le x) = \frac{1}{1 + \exp(-x)}$$

:::

## Latent variables

::: {.incremental}

- So what's $\mathbb{E}[Y_i|X_i] = \Pr(Y_i = 1|X_i)$?

  $$\Pr(Y_i = 1|X_i) = \Pr(Y_i^* \ge 0 | X_i) = \Pr(\epsilon_i \ge -X_i^{\prime}\beta) = \Pr(\epsilon_i \le X_i^{\prime}\beta) = \frac{1}{1 + \exp(-X_i^{\prime}\beta)}$$

  - The last-but-one step uses the **symmetry** of the logistic distribution around zero.
- We can write the **probit** similarly, defining the error distribution as

  $$\epsilon_i \sim \text{Normal}(0, 1)$$

- And by extension, the probability $\Pr(Y_i = 1|X_i)$

  $$\Pr(Y_i = 1|X_i) = \Pr(Y_i^* \ge 0|X_i) = \Pr(\epsilon_i \le X_i^{\prime}\beta) = \Phi(X_i^{\prime}\beta)$$

- **Note** the scale of $\epsilon_i$ is fixed by assumption, not estimated
  - $Y_i^*$ is never observed, so only the *ratio* $\beta/s$ is identified -- so we normalize to $s =1$

:::

## Latent variables

- We can visualize this as an integral over the density of the random "error" $\epsilon_i$

```{r, echo=F, fig.width=5, fig.height = 3, fig.align="center"}
logistic_dens <- ggplot() + xlim(-5, 5) + geom_function(fun=dlogis) + theme_bw() + ylab("Density") + geom_area(aes(x=c(-5,2)),stat = "function", fun = dlogis, fill = "dodgerblue", xlim = c(-5, 2)) + 
  xlab(expression(epsilon)) + geom_vline(xintercept=0, lty=2) + geom_vline(xintercept=2, lty=3, col="indianred", lwd=1) 

logistic_dens
```

## Latent variables

- Same thing for a probit

```{r, echo=F, fig.width=5, fig.height = 3, fig.align="center"}
probit_dens <- ggplot() + xlim(-5, 5) + geom_function(fun=dnorm) + theme_bw() + ylab("Density") + geom_area(aes(x=c(-5,2)),stat = "function", fun = dnorm, fill = "dodgerblue", xlim = c(-5, 2)) + 
  xlab(expression(epsilon)) + geom_vline(xintercept=0, lty=2) + geom_vline(xintercept=2, lty=3, col="indianred", lwd=1) 

probit_dens
```

## Ordinal logit

::: {.incremental}

- Using the latent variable formulation lets us extend binary outcome models to outcomes with multiple discrete levels.
- For example, consider an "ordinal" response variable
  - (e.g.) a Likert scale (1 = "strongly disagree", 3 = "neutral", 5 = "strongly agree")
- The responses are *ordered* but the intervals between them do not have any meaning.
  - A popular modeling technique with an ordinal $Y$ is to treat it as an expression of a latent variable.

:::

## Ordinal logit

::: {.incremental}

- In an *ordered logit* model, an ordinal outcome $Y$ with $L$ unique, ordered values $Y \in \{1, 2, 3, \dotsc L\}$ is a function of a latent variable and a set of $L-1$ cutpoints $\kappa$

  $$Y_i = \begin{cases}1 &\text{if } Y_i^* \le \kappa_1 \\
                        2 &\text{if } \kappa_1 < Y_i^* \le \kappa_2\\
                        3 &\text{if } \kappa_2 < Y_i^* \le \kappa_3\\
                        \vdots &\\
                        L &\text{if } Y_i^* > \kappa_{L-1}\end{cases}$$

- We still assume $Y_i^*$ has a logistic distribution

  $$Y_i^* = X_i^{\prime}\beta + \epsilon_i, \qquad \epsilon_i \sim \text{Logistic}(0, 1)$$

:::

## Ordinal logit

::: {.incremental}

- To estimate the model, we now have to estimate both the $\beta$ parameters and the $\kappa$ parameters
  - Note we omit the intercept from $X_i$ -- the $\kappa$ parameters are essentially the "intercepts" for each boundary between the choices.
- How does the model map changes in the latent variable to changes in probability? Work with the **cumulative** probabilities:

  $$\Pr(Y_i \le l \,|\, X_i) = \Pr(Y_i^* \le \kappa_l \,|\, X_i) = \Pr(\epsilon_i \le \kappa_l - X_i^{\prime}\beta)$$

- And then difference adjacent cumulatives to get each cell probability

  $$\Pr(Y_i = l \,|\, X_i) = \Pr(Y_i \le l \,|\, X_i) - \Pr(Y_i \le l-1 \,|\, X_i)$$

  with the conventions $\Pr(Y_i \le 0) = 0$ and $\Pr(Y_i \le L) = 1$.

:::

## Ordinal logit

- Let's think in terms of the distribution of $Y_i^*$ and the cut-points for a case with $L=3$. For $X_i^{\prime}\beta = 0$:

```{r, echo=F, fig.width=7, fig.height = 3, fig.align="center"}
logistic_dens <- ggplot() + xlim(-8, 8) + geom_function(fun=dlogis) + theme_bw() + ylab("Density") + geom_area(aes(x=c(-8,-1)),stat = "function", fun = dlogis, fill = "dodgerblue", xlim = c(-8, -1)) +  geom_area(aes(x=c(-1,2)),stat = "function", fun = dlogis, fill = "darkorange", xlim = c(-1, 2)) + geom_area(aes(x=c(2,8)),stat = "function", fun = dlogis, fill = "gold", xlim = c(2, 8)) + 
  xlab(expression(Y^"*")) + geom_vline(xintercept=0, lty=2) + geom_vline(xintercept=-1, lty=3, col="darkgreen", lwd=1) + geom_vline(xintercept=2, lty=3, col="darkgreen", lwd=1) 

logistic_dens
```

- The green lines are the cutpoints $\kappa_1, \kappa_2$; the shaded areas are $\Pr(Y_i = 1), \Pr(Y_i = 2), \Pr(Y_i = 3)$.

## Ordinal logit

- Shifting the mean of $Y^*$ up by 3:

```{r, echo=F, fig.width=7, fig.height = 3, fig.align="center"}
logistic_dens <- ggplot() + xlim(-8, 8) + geom_function(fun=dlogis, args=list(location=3)) + theme_bw() + ylab("Density") + geom_area(aes(x=c(-8,-1)),stat = "function", fun = dlogis, fill = "dodgerblue", xlim = c(-8, -1),  args=list(location=3)) +  geom_area(aes(x=c(-1,2)),stat = "function", fun = dlogis, fill = "darkorange", xlim = c(-1, 2),  args=list(location=3)) + geom_area(aes(x=c(2,8)),stat = "function", fun = dlogis, fill = "gold", xlim = c(2, 8),  args=list(location=3)) + 
  xlab(expression(Y^"*")) + geom_vline(xintercept=3, lty=2) + geom_vline(xintercept=-1, lty=3, col="darkgreen", lwd=1) + geom_vline(xintercept=2, lty=3, col="darkgreen", lwd=1) 

logistic_dens
```

- The cutpoints stay put but the *distribution* slides. All three probabilities move together as a result.

## Ordinal logit

::: {.incremental}

- This structural mapping from $Y^*$ to $\Pr(Y_i = l)$ encodes a particular assumption about how these probabilities relate to one another.
- Referred to as the **proportional odds** assumption

  $$\log\left(\frac{\Pr(Y_i \le l)}{1 - \Pr(Y_i \le l)}\right) = \kappa_l - X_i^{\prime}\beta$$

  - The same $\beta$ appears for *every* $l$; only the intercept $\kappa_l$ changes.
- **Intuitively**: the $\beta$ parameters are shared across all the possible outcome levels. 
  - Information about whether someone responds "strongly agree" *also* tells us something about how likely they would be to respond "agree", "neutral", "disagree", or "strongly disagree".
  - Not innocuous if there's non-monotonicity or heterogeneous effects.
  - Be careful with these models!

:::

## Multinomial logit

::: {.incremental}

- In a setting with multiple **unordered** outcomes, the natural extension of the logit regression is the **multinomial** logit
- Suppose we have an outcome with $K$ unordered levels (e.g. "choices"). Define a latent "utility" $Y_{ik}^*$ for each of the $K$ categories
  - For identification, we set one category to be the "baseline" by anchoring its coefficients to zero: $\beta_1 = 0$

  $$\begin{align*}Y_{i1}^* &= X_i^{\prime}\beta_1 + \epsilon_{i1} = \epsilon_{i1}\\
  Y_{i2}^* &= X_i^{\prime}\beta_2 + \epsilon_{i2}\\
  &\vdots \\
  Y_{iK}^* &= X_i^{\prime}\beta_K + \epsilon_{iK}\end{align*}$$

- The probability of observing choice $Y_{i} = k$ is the probability that the $k$th latent variable is greater than all of the others

  $$\Pr(Y_i = k \,|\, X_i) = \Pr\big(Y^*_{ik} = \max\{Y^*_{i1}, Y^*_{i2}, \dotsc Y^*_{iK}\}\big)$$

:::

## Multinomial logit

::: {.incremental}

- It turns out that if the error terms $\epsilon_{ik}$ have i.i.d. "Type I extreme value" distributions, we can write the choice probabilities in closed form

  $$\Pr(Y_i = k \,|\, X_i) = \frac{\exp(X_i^{\prime}\beta_{k})}{\sum_{l=1}^K \exp(X_i^{\prime}\beta_{l})}$$

- This is sometimes referred to as the "softmax" function in machine learning.
- **Note**: with the normalized baseline category $\beta_1 = 0$, we can write this equivalently as

  $$\Pr(Y_i = k \,|\, X_i) = \frac{\exp(X_i^{\prime}\beta_{k})}{1 + \sum_{l=2}^K \exp(X_i^{\prime}\beta_{l})}$$

- The i.i.d. error assumption implies **independence of irrelevant alternatives** (IIA).
  - The relative odds of choice $k$ vs. $l$ don't depend on what *other* options are available.
  - Famously implausible in some choice settings (the "red bus/blue bus" problem).

:::

## Example: Appointer Affiliation Bias

- @puig2017affiliation ran a survey experiment on a group of international arbitration experts to see how manipulating the appointing party affected their decisionmaking in a vignette
  - Respondents were given a hypothetical dispute and asked to determine who should pay the legal costs
  - Three options: **Loser pays all**, **Loser pays some**, **Both sides split**

## Example: Appointer Affiliation Bias

- Read in the responses -- `-99` is the survey platform's missing-value code

```{r, echo=T, message = F}
study1 <- read_csv("data/experiment1_cleaned.csv", na = c("", "NA", "-99"))
```

- The outcome is stored as an integer, so recode it into an **ordered factor**
  - The ordering is what makes this an ordinal outcome rather than a nominal one

```{r, echo=T}
study1 <- study1 %>%
  mutate(isReimbursementFolded_Factor = case_when(
    isReimbursementFolded == 1 ~ "Winner pays all",
    isReimbursementFolded == 2 ~ "Winner pays some",
    isReimbursementFolded == 3 ~ "Split costs",
    isReimbursementFolded == 4 ~ "Loser pays some",
    isReimbursementFolded == 5 ~ "Loser pays all"))

study1$isReimbursementFolded_Factor <- factor(
  study1$isReimbursementFolded_Factor,
  levels = c("Winner pays all", "Winner pays some", "Split costs",
             "Loser pays some", "Loser pays all"), ordered = TRUE)
```

## Example: Appointer Affiliation Bias

- Keep the completed responses, and drop the two "winner pays" categories
  - Essentially no one chose them, so we're left with the **three** options in the vignette

```{r, echo=T}
study1_complete <- study1 %>%
  filter(!is.na(isReimbursementFolded) & Finished == 1) %>%
  filter(isReimbursementFolded_Factor != "Winner pays some" &
         isReimbursementFolded_Factor != "Winner pays all")

table(study1_complete$isReimbursementFolded_Factor)
```

## Example: Appointer Affiliation Bias

```{r, echo=F, fig.align="center", fig.width=7, fig.height=3.2}
study1_complete %>% ggplot(aes(x=isReimbursementFolded_Factor)) + geom_bar() + xlab("Selected outcome") + ylab("Count") + ggtitle("Distribution of outcome in Puig and Strezhnev (2017)") + theme_minimal()
```

## Example: Appointer Affiliation Bias

- It would seem like an ordinal logit regression would be appropriate here -- a three-category discrete outcome with a clear ordering
  - Implemented in the `ordinal` package

```{r, echo=T}
ordered_logit <- ordinal::clm(isReimbursementFolded_Factor ~ isAppointerSide, data=study1_complete %>% 
                       filter(isAppointerSide == "Appointed by Winner"|isAppointerSide == "Appointed by Loser"))
```

```{r, echo=T}
summary(ordered_logit)
```

## Example: Appointer Affiliation Bias

- The model implies that arbitrators appointed by the winning party shift mostly from **splitting the costs** to making the loser pay for **everything**
  - But is this real or an artifact of the proportional odds assumption? Is there a "constant effect" on the latent variable scale?

```{r, echo=T}
ord_me <- marginaleffects::avg_comparisons(ordered_logit)
```

```{r, echo=F}
# Bar plot of the change in each category's probability (winner- vs. loser-appointed), with 95% CIs
plot_me <- function(me) {
  me %>%
    mutate(group = factor(group, levels = c("Split costs", "Loser pays some", "Loser pays all"))) %>%
    ggplot(aes(x = group, y = estimate, fill = model)) +
    geom_hline(yintercept = 0, color = "grey40") +
    geom_col(position = position_dodge(width = 0.75), width = 0.7) +
    geom_errorbar(aes(ymin = conf.low, ymax = conf.high), position = position_dodge(width = 0.75), width = 0.15) +
    scale_fill_manual(values = c("Ordinal logit" = "#c5050c", "Multinomial logit" = "#0479A8"), name = NULL) +
    labs(x = NULL, y = "Change in Pr(category)
winner- vs. loser-appointed") +
    theme_minimal(base_size = 13) +
    theme(legend.position = "top", panel.grid.major.x = element_blank())
}
```

```{r, echo=F, fig.width=7, fig.height=2.6, fig.align="center"}
plot_me(as_tibble(ord_me) %>% mutate(model = "Ordinal logit"))
```

## Example: Appointer Affiliation Bias

- Compare with **multinomial logit** (which in this case is equivalent to OLS for each category since we're using a fully-saturated regression)

```{r, echo = T, warning = F, message = F, results = "hide"}
multinom_logit <- nnet::multinom(isReimbursementFolded_Factor ~ isAppointerSide, data=study1_complete %>% 
                       filter(isAppointerSide == "Appointed by Winner"|isAppointerSide == "Appointed by Loser"))
```

```{r, echo=T}
mnl_me <- marginaleffects::avg_comparisons(multinom_logit)
```

```{r, echo=F, fig.width=7, fig.height=3, fig.align="center"}
bind_rows(as_tibble(ord_me) %>% mutate(model = "Ordinal logit"),
          as_tibble(mnl_me) %>% mutate(model = "Multinomial logit")) %>%
  mutate(model = factor(model, levels = c("Ordinal logit", "Multinomial logit"))) %>%
  plot_me()
```

- Running the ordinal logit (as a reviewer asked us to do!) would have given us the **wrong result**
  - Remember, don't just fixate on the outcome when choosing a modeling strategy - think about what your entire bundle of assumptions implies!

## Conditional logit

::: {.incremental}

- Same latent utility setup as the multinomial logit, but now the covariates are **attribute-specific** but their types are **shared** across alternatives
  - Price of each mode of transportation, the attributes of each profile in a conjoint
- $X_{ik}$ varies across categories, but all categories share a common $\beta$ vector

  $$Y_{ik}^* = X_{ik}^{\prime}\beta + \epsilon_{ik}$$

- With the same i.i.d. Type I extreme value errors, we get the same "softmax" expression for the choice probability.

  $$\Pr(Y_i = k \,|\, X_{i1}, \dotsc, X_{iK}) = \frac{\exp(X_{ik}^{\prime}\beta)}{\sum_{l=1}^K \exp(X_{il}^{\prime}\beta)}$$

:::


## IIA

::: {.incremental}

- We still have the independence of irrelevant alternatives assumption: the relative odds of two choices only depend on those two choices' attributes.

  $$\frac{\Pr(Y_i = j)}{\Pr(Y_i = k)} = \exp\big((X_{ij} - X_{ik})^{\prime}\beta\big)$$

- Extensions involve imposing additional **hierarchical** structure:
  - **Nested logit** - group similar alternatives and allow correlation within groupings 
    - A random effect common to every alternative in the same **group** $g(k)$: $Y_{ik}^* = X_{ik}^{\prime}\beta + \zeta_{i,g(k)} + \epsilon_{ik}$
  - **Mixed logit / random coefficients** -- let $\beta$ vary across individuals [@train2009discrete]
    - **Unit-specific** coefficients drawn from a common distribution: $Y_{ik}^* = X_{ik}^{\prime}\beta_i + \epsilon_{ik}$
  - Increasingly estimated via **Bayesian** methods (for regularization + partial pooling)

:::

## Conjoint experiments

::: {.incremental}

- A **conjoint** experiment shows respondents two (or more) profiles with **randomly assigned** attributes and asks them to pick one.
  - e.g. choosing between two candidates with different backgrounds, partisan affiliation, policy positions, etc...
- Often designed as a forced choice between alternatives described by **alternative-specific** covariates 
- In **political science**: analyzed as a factorial experiment
  - Estimand is the **AMCE** [@hainmueller2014causal] -- the average change in the probability of selection from changing one attribute, marginalizing over the others.
  - Estimated by OLS, no distributional assumptions, "causal" interpretation.
- In literally every other field that does conjoints: analyzed with a **random utility model**
  - Model latent utilities and simulate choice probabilities.

:::

## Example: Workplace democracy

::: {.incremental}

- @mazumder2024workplace ask whether Americans want to work at **democratically governed** firms
  - YouGov Blue, June 2020: 1,002 respondents, four tasks each, pairwise comparison of two hypothetical employers
  - Outcome: at which firm would you **rather work**? (forced choice)
- Each firm profile randomizes **19 attributes**: the one emphasized in the paper is **corporate governance**
  - *Private, non-worker shareholders* (baseline), *Private, one owner*, *Public, shareholders*, *Workers are shareholders*, *Workers on the board*, *Workers elect managers*
  - Plus pay and benefits, hours, location, owners' race and gender, unionization, political donations, workplace conditions, ...
- The paper reports **AMCEs** from OLS. We'll fit the **conditional logit** to the same choices.

:::

## Example: Workplace democracy

- One row per respondent-task-**profile**: exactly the long format the conditional logit wants
  - We drop the 122 tasks where the respondent picked neither firm, leaving 3,886 tasks from 993 respondents

```{r, echo=T}
library(mlogit)
library(sandwich)
cj <- readRDS("data/mazumder_yan_conjoint.rds")
cj %>% select(rid, task, profile, chosen, governance, income, hours) %>% head(4)
```

- `dfidx()` tells `mlogit` which rows form a choice situation (`task_id`, nested in respondent `rid`) and which alternative each row is (`profile`)

```{r, echo=T}
cj <- cj %>% mutate(task_id = paste(rid, task, sep = "_"))
cj_idx <- dfidx(cj, idx = list(c("task_id", "rid"), "profile"))
```

## Example: Workplace democracy

- All 19 attributes are **alternative-specific** with a **generic** coefficient: the first part of the `mlogit` formula
  - `| 0` drops the alternative-specific constant (a preference for whichever profile is listed second)

```{r, echo=T}
attributes <- names(cj)[5:23]
cl_formula <- as.formula(paste("chosen ~", paste(attributes, collapse = " + "), "| 0"))
cl_fit <- mlogit(cl_formula, data = cj_idx)
length(coef(cl_fit))
```

- Respondents complete four tasks each, so cluster by respondent
  - `estfun()` has one row per **task** (one likelihood contribution each), so the cluster vector needs one entry per task

```{r, echo=T}
task_rid <- idx(cj_idx)$rid[idx(cj_idx)$profile == 1]
V_cl <- vcovCL(cl_fit, cluster = task_rid)
```

## Example: Workplace democracy

- The coefficients are **part-worths**: each level's contribution to the systematic component of utility, relative to its attribute's baseline
- For two firms that are **identical except for governance**, $\Pr(\text{choose } A) = \text{logit}^{-1}(\beta_{A} - \beta_{B})$

```{r, echo=T}
gov <- grep("^governance", names(coef(cl_fit)), value = TRUE)
tibble(level = sub("governance", "", gov), part_worth = coef(cl_fit)[gov],
       se = sqrt(diag(V_cl))[gov], odds_ratio = exp(part_worth),
       pr_vs_baseline = plogis(part_worth)) %>% kable(digits = 3)
```

- Codetermination: a firm with workers on the board beats an otherwise identical private corporation about **`r round(100 * plogis(coef(cl_fit)["governanceWorkers on the board"]))`%** of the time

```{r, echo = F}
# Part-worths for every level, baselines included at zero, for the plots below
attr_labels <- c(governance = "Corporate governance", csr = "Social responsibility", firm_size = "Firm size",
                 donations = "Political contributions", gender_owned = "Gender of owners", race_owned = "Race of owners",
                 union = "Unionization", training = "Job training", income = "Salary",
                 healthcare = "Health insurance\n(employer pays)", retirement = "401k matching",
                 sick_leave = "Paid sick leave", parental_leave = "Parental leave", hours = "Hours per week",
                 location = "Location", remote = "Work from home", work_type = "Type of task",
                 teamwork = "Work with others", culture = "Workplace\nconditions")
part_worths <- map_dfr(attributes, function(a) {
  lv <- levels(cj[[a]]); term <- paste0(a, lv)
  tibble(attribute = a, label = attr_labels[[a]], level = lv, baseline = seq_along(lv) == 1,
         estimate = ifelse(baseline, 0, coef(cl_fit)[term]),
         se = ifelse(baseline, 0, sqrt(diag(V_cl))[term]))
}) %>% mutate(conf.low = estimate - 1.96 * se, conf.high = estimate + 1.96 * se)

plot_part_worths <- function(attrs) {
  part_worths %>%
    filter(attribute %in% attrs) %>%
    # Levels repeat across attributes ("None", "Mix"), so key each row on
    # attribute + level and strip the attribute only for the axis labels
    mutate(label = factor(label, levels = attr_labels[attrs]),
           row = factor(paste(attribute, level), levels = rev(paste(attribute, level)))) %>%
    ggplot(aes(x = estimate, y = row)) +
    scale_y_discrete(labels = function(x) sub("^\\S+ ", "", x)) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
    geom_errorbar(aes(xmin = conf.low, xmax = conf.high), width = 0, color = "#c5050c", linewidth = 0.7) +
    geom_point(aes(shape = baseline), color = "#c5050c", fill = "white", size = 2) +
    scale_shape_manual(values = c(`FALSE` = 19, `TRUE` = 21), guide = "none") +
    facet_grid(label ~ ., scales = "free_y", space = "free_y", switch = "y") +
    labs(x = "Part-worth (log-odds, relative to baseline) with 95% CI", y = NULL) +
    theme_minimal(base_size = 11) +
    theme(strip.placement = "outside", strip.text.y.left = element_text(angle = 0, hjust = 1, face = "bold"),
          panel.spacing = unit(2, "pt"), panel.grid.minor = element_blank())
}
```

## Part-worths: the firm

```{r, echo = F, fig.width = 10, fig.height = 5.6, fig.align = "center"}
plot_part_worths(c("governance", "csr", "firm_size", "donations", "gender_owned", "race_owned", "union", "training"))
```

::: {.attribution}
Conditional logit with all 19 attributes; hollow points are baselines. 95% CIs clustered by respondent. Data from @mazumder2024workplace.
:::

## Part-worths: the job

```{r, echo = F, fig.width = 10, fig.height = 5.6, fig.align = "center"}
plot_part_worths(c("healthcare", "retirement", "sick_leave", "parental_leave", "hours", "location", "remote", "work_type", "teamwork", "culture"))
```

::: {.attribution}
Conditional logit with all 19 attributes; hollow points are baselines. 95% CIs clustered by respondent. Data from @mazumder2024workplace.
:::

## Part-worths: salary

- 28 salary levels, so each dummy is noisy. A single **log-salary** term seems to fit well:

```{r, echo=T}
cj <- cj %>% mutate(log_salary = log(as.numeric(gsub("[^0-9]", "", income))))
log_formula <- as.formula(paste("chosen ~", paste(c(setdiff(attributes, "income"), "log_salary"), collapse = " + "), "| 0"))
log_fit <- mlogit(log_formula, data = dfidx(cj, idx = list(c("task_id", "rid"), "profile")))
lr_salary <- as.numeric(2 * (logLik(cl_fit) - logLik(log_fit)))   # 27 dummies vs. 1 log term
c(LR = lr_salary, df = length(coef(cl_fit)) - length(coef(log_fit)),
  p = pchisq(lr_salary, length(coef(cl_fit)) - length(coef(log_fit)), lower.tail = FALSE))
```

```{r, echo = F, fig.width = 10, fig.height = 2.9, fig.align = "center"}
V_log <- vcovCL(log_fit, cluster = task_rid)
b_log <- coef(log_fit)["log_salary"]; se_log <- sqrt(V_log["log_salary", "log_salary"])
# Part-worths are only identified up to a constant within an attribute, so center the
# log-salary curve on the dummies' average level (and pivot its CI at the mean log salary)
salary_pw <- part_worths %>% filter(attribute == "income") %>%
  mutate(salary = as.numeric(gsub("[^0-9]", "", level)))
center <- mean(log(salary_pw$salary)); level0 <- mean(salary_pw$estimate)
salary_curve <- tibble(salary = seq(30, 300, 1)) %>%
  mutate(fit = level0 + b_log * (log(salary) - center),
         lo = level0 + (b_log - 1.96 * se_log * sign(log(salary) - center)) * (log(salary) - center),
         hi = level0 + (b_log + 1.96 * se_log * sign(log(salary) - center)) * (log(salary) - center))
salary_pw %>%
  ggplot(aes(x = salary)) +
  geom_ribbon(data = salary_curve, aes(ymin = lo, ymax = hi), fill = "#0479A8", alpha = 0.15) +
  geom_line(data = salary_curve, aes(y = fit), color = "#0479A8", linewidth = 1) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0, color = "#c5050c", alpha = 0.6) +
  geom_point(aes(y = estimate, shape = baseline), color = "#c5050c", fill = "white", size = 2) +
  scale_shape_manual(values = c(`FALSE` = 19, `TRUE` = 21), guide = "none") +
  scale_x_continuous(labels = function(x) paste0("$", x, "k"), breaks = seq(50, 300, 50)) +
  labs(x = "Salary offer", y = "Part-worth\n(relative to $30k)") +
  theme_minimal(base_size = 12)
```

::: {.attribution}
Red: salary dummies with 95% CIs. Blue: log-salary fit with 95% CI, centered on the dummies' average level. Both clustered by respondent.
:::

## Predicted choices {.app-slide}

::: {.app-frame data-app="clogit_app.html" data-title="Conditional logit predictions: Mazumder and Yan (2024)"}
:::

::: {.attribution}
Conditional logit part-worths and respondent-clustered covariance from @mazumder2024workplace. CIs: delta method on the log-odds scale.
:::

```{=html}
<script>
// The widget is a separate page framed here. The <iframe> is built at runtime
// because `embed-resources: true` would otherwise inline the target into a
// data: URI, breaking the app's relative asset paths and its service worker.
// Mounting only once the slide is reached also delays the webR download.
(function () {
  function mount() {
    document.querySelectorAll("section.present .app-frame[data-app]").forEach(function (el) {
      if (el.dataset.mounted) return;
      el.dataset.mounted = "1";
      var f = document.createElement("iframe");
      f.src = el.dataset.app;
      f.title = el.dataset.title || "Interactive widget";
      f.setAttribute("allow", "clipboard-write");
      el.appendChild(f);
    });
  }
  // Reveal is not defined while this script is parsed, so wire up after load.
  window.addEventListener("load", function () {
    if (window.Reveal && typeof Reveal.on === "function") Reveal.on("slidechanged", mount);
    mount();
  });
})();
</script>
```

## Survival models

## Survival models


::: {.incremental}

- Often we have outcome data that is measured as the "time-to-event"
  - From the medical literature, this is often the number of days before a patient suffers some adverse event (e.g. death).
  - But lots of other applications (e.g. time to treaty ratification)

- Assume we observe $n$ i.i.d. event times $T_1, T_2, \dotsc, T_n$

- Define the "survival function"

  $$S(t) = Pr(T > t) = 1 - F(t)$$

- We'll also often work with the "hazard" or the "instantaneous" probability of an event happening *given* that it has not yet occurred

  $$h(t) = \underset{\Delta t \to 0}{\lim} \frac{P(t < T_i \le t + \Delta t | T_i > t)}{\Delta t} = \frac{f(t)}{S(t)} = \frac{f(t)}{1-F(t)}$$

- **Intuitively**: Given that I know that an event has not occurred until $t$, what is the probability that it will immediately occur next!

:::

## Survival models

- We'll also define the cumulative hazard

  $$H(t) = \int_{0}^{t} h(u)du = \int_0^{t} \frac{f(u)}{S(u)} du =  -\log(S(t))$$

## Non-parametric approach

::: {.incremental}

- The easiest way to estimate $S(t) = 1 - F(t)$ is to just plug in $\hat{F}(t)$ - our empirical CDF in the sample
  - At any given value $t$, what share of events occur *after* $t$

- **Problem**: This doesn't address *censoring* of observations -- sometimes we stop observing a unit at time $t^*$ rather than observing an actual event.
  - We know that $T_i > t^*$ but not what $T_i$ actually equals
  
- Let $C = \{C_1, C_2, \dotsc, C_n\}$ denote the **censoring indicator**
  - If $C_i = 1$, then the observation is *censored*, the observed value $t$ only gives us information on $T_i > t$ as opposed to $T_i = t$.
  - Crucially, we assume that censoring is "uninformative"

- One approach to constructing a non-parametric estimator of the survival function *under censoring* is the **Kaplan-Meier** estimator.

:::

## Kaplan-Meier Estimator

::: {.incremental}

- Start by arranging the possible event times from $0 < t_{(1)} < t_{(2)} < t_{(3)} < t_{(4)}, \dotsc, < t_{(n)}$
  - Suppose $t$ is after $t_{(j)}$ and before $t_{(j+1)}$. We can write the survival function in terms of conditional probabilities:
  
  $$S(t) = Pr(T_i > t | T_i > t_{(j)}) \times Pr(T_i > t_{(j)} | T_i > t_{(j-1)}) \times \dotsc \times Pr(T_i > t_{(2)} | T_i  > t_{(1)}) \times Pr(T_i > t_{(1)})$$

- The **Kaplan-Meier** estimator $\hat{S}(t)$ can be understood as a "plug-in" estimator for each of these probabilities.

  $$\hat{S}(t) = \prod_{j: t_{(j)} \le t} \bigg(1 - \frac{d_j}{r_j} \bigg)$$

- $d_j$ denotes the number of units that experience an event at time $t_{(j)}$ 
- $r_j$ is the number of units that are still in the risk set just before $t_{(j)}$
  - Units drop out of the risk set $r$ over time by either experiencing an event or being censored

:::

## Kaplan-Meier Estimator

::: {.incremental}

- **Intuition**: Consider one of the terms in the product. Using the definition of conditional probability:

  $$Pr(T_i > t_{(j)} | T_i > t_{(j-1)}) = 1 - Pr(T_i \le t_{(j)}|T_i > t_{(j-1)}) = 1 - \frac{Pr(T_i \le t_{(j)}, T_i > t_{(j-1)})}{Pr(T_i > t_{(j-1)})}$$

- In the numerator, our "plug-in" estimator is just the share of units that have events occurring at $t_{(j)}$.
- In the denominator we have the share of units that we know for sure are still "active" in the interval after $t_{(j-1)}$

- Hence our plug-in estimator:

  $$\widehat{Pr(T_i > t_{(j)} | T_i > t_{(j-1)})} = 1 - \frac{d_j}{r_j}$$

- Taking the product across all of the $j$ where $t_{(j)} \le t$ yields the Kaplan-Meier estimator.

:::

## Illustration: Lung Cancer

- Let's look at a sample dataset from the `survival` R package on survival times for patients with advanced lung cancer.

```{r, echo=T}
library(survival)
data(cancer, package="survival")
as_tibble(lung)
```

- We want to estimate $S(t | \text{Sex} = \text{Male})$ and $S(t | \text{Sex} = \text{Female})$


## Illustration: Lung Cancer

- In the `survival` package, we need to define a "survival" outcome in terms of the event time and the censoring indicator

```{r, echo=T}
surv_model <- survfit(Surv(time, status) ~ 1, data=lung)
```



## Illustration: Lung Cancer

- Plot the curve for the sample as a whole

```{r, fig.width=8, fig.height=6, fig.align='center'}
plot(surv_model, xlab="Survival Time", conf.int=T, ylab="Survival Rate", mark.time=TRUE)
```



## Illustration: Lung Cancer

- Now let's do this for gender

```{r, echo=T}
surv_model_sex <- survfit(Surv(time, status) ~ I(sex==2), data=lung)
```



## Illustration: Lung Cancer

- Plot separate curves for men and women

```{r, fig.width=8, fig.height=6, fig.align='center'}
plot(surv_model_sex, col=c("indianred", "dodgerblue"), xlab="Survival Time", conf.int=T, ylab="Survival Rate", mark.time=T)
legend(750, .9, c("Male", "Female"), col=c("indianred", "dodgerblue"), lwd=2, bty='n')
```



## Parametric inference

::: {.incremental}

- When we want to model the survival function using covariates, we run into a clear constraint using purely non-parametric methods -- with continuous covariates, we may only have one observation that has a unique covariate value
  - Splitting the sample and constructing separate K-M curves quickly becomes infeasible.

- We may want to assume a **model** for the event time. We can write that density in terms of the *hazard* and the *survival* function

  $$f(t) = h(t)S(t)$$

- Note that by the definition of the cumulative hazard $H(t) = - \log(S(t))$ and $S(t) = \exp(-H(t))$, so 

  $$f(t) = h(t)\exp(-H(t))$$

:::

## Parametric inference

- Depending on the assumptions we make about the nature of the hazard, we can obtain different distributions for $f(t)$.
  - In the simplest case, let's assume that the hazard does not vary over time: $h(t) = \lambda$, with $\lambda > 0$
  - And the cumulative hazard is $H(t) = \int_{0}^{t} \lambda du = \lambda t$


- This yields the **exponential distribution**

  $$f(t) = \lambda \exp(-\lambda t)$$


## Exponential Model

- The mean of $T_i \sim \text{Exponential}(\lambda)$ is $E[T_i] = \frac{1}{\lambda}$. We'll often reparameterize the mean as $\theta$, yielding a density of 

  $$f(t) = \frac{1}{\theta} \exp\bigg(-\frac{1}{\theta} t\bigg)$$

- $\theta$ is strictly greater than $0$, so we'll use a log-link to incorporate the covariates:

  $$\log(E[T_i | X_i]) = \log(\theta_i) = X_i^{\prime}\beta$$

- Alternatively, we can write 

  $$E[T_i | X_i] = \exp(X_i^{\prime}\beta)$$

- Note that the constant hazard assumption is a strong one -- the exponential distribution is "memoryless" in that knowing that a unit has survived up until time $t$ tells you nothing about its instantaneous probability of failure.
  - More flexible distributions like the Weibull allow for a time-varying hazard.

## Exponential Model

 - We can interpret changes in the linear predictor in terms of changes in the hazard since 
 
 $$h(t) = \frac{1}{\theta_i} = \frac{1}{\exp(X_i^{\prime}\beta)}$$
 
 - Increases in the linear predictor reduce the hazard.
 
 - This also lets us write down the survival function
 
 $$S(t) = \exp\bigg(- \frac{1}{\exp(X_i^{\prime}\beta)}t \bigg)$$

## Application

- Let's compare our parametric fit to the non-parametric Kaplan-Meier estimates
 
```{r, echo=T}
# sex = 2 is a dummy variable for female
exp_model_sex <- survreg(Surv(time, status) ~ I(sex==2), dist="exponential", data=lung)
```

- Get the survival curves for men and women

```{r, echo=T}
exp_surv_men <- function(x) pexp(x, rate = 1/exp(coef(exp_model_sex)%*%c(1,0)), lower.tail = F)
exp_surv_women <- function(x) pexp(x, rate = 1/exp(coef(exp_model_sex)%*%c(1,1)), lower.tail=F)
```


## Application

- Overlay the exponential survival curves on the Kaplan-Meier estimates

```{r, fig.width=8, fig.height=6, fig.align='center', warning=F}
plot(surv_model_sex, col=c("indianred", "dodgerblue"), xlab="Survival Time", conf.int=T, ylab="Survival Rate", mark.time=T)
legend(750, .9, c("Male", "Female"), col=c("indianred", "dodgerblue"), lwd=2, bty='n')
curve(exp_surv_men, from = 0, to = 1000, lwd=2, col="indianred", add=T)
curve(exp_surv_women, from = 0, to = 1000, lwd=2, col="dodgerblue", add=T)
```


## Application: Weibull

::: {.incremental}

- The fit seems to be off.
  - This is because the hazard appears to be **time-varying**.
  - You can see a kind of "s-curve" which the exponential cannot capture.
- The **Weibull** adds a shape parameter $k$ that lets the hazard change over time

  $$h(t) = \frac{k}{\lambda}\left(\frac{t}{\lambda}\right)^{k-1}\qquad S(t) = \exp\left(-\left(\frac{t}{\lambda}\right)^{k}\right)$$

  - $k = 1$ is the exponential; $k > 1$ is a **rising** hazard, $k < 1$ a falling one

:::

```{r, echo=T}
weib_model_sex <- survreg(Surv(time, status) ~ I(sex==2), dist="weibull", data=lung)
```

## Application: Weibull

- Get the Weibull survival curves ($\lambda_i = \exp(X_i^{\prime}\beta)$) and overlay them on Kaplan-Meier

```{r, echo=T}
weib_surv_men <- function(x) pweibull(x, shape = 1/weib_model_sex$scale,
                                       scale = exp(coef(weib_model_sex)%*%c(1,0)), lower.tail = F)
weib_surv_women <- function(x) pweibull(x, shape = 1/weib_model_sex$scale,
                                         scale = exp(coef(weib_model_sex)%*%c(1,1)), lower.tail = F)
```

```{r, fig.width=8, fig.height=4.3, fig.align='center', warning=F}
plot(surv_model_sex, col=c("indianred", "dodgerblue"), xlab="Survival Time", conf.int=T, ylab="Survival Rate", mark.time=T)
legend(750, .9, c("Male", "Female"), col=c("indianred", "dodgerblue"), lwd=2, bty='n')
curve(weib_surv_men, from = 0, to = 1000, lwd=2, col="indianred", add=T)
curve(weib_surv_women, from = 0, to = 1000, lwd=2, col="dodgerblue", add=T)
```


## Cox Proportional Hazards

- The **Cox proportional hazards model** assumes that covariates enter the hazard log-linearly, but the baseline hazard is unspecified

$$h_i(t) =  h_0(t)\exp(X_i^{\prime}\beta)$$

- $h_0(t)$ is the unknown baseline hazard function.
- Note that while the baseline hazard is not explicitly identified, the ratio of the hazards of two observations $i$ and $j$ at time $t$ **does not depend on the baseline**

  $$\frac{h_i(t)}{h_j(t)} = \frac{\exp(X_i^{\prime}\beta)}{\exp(X_j^{\prime}\beta)}$$

- The model is **semi-parametric** in that some components are left entirely unspecified...
  - ...but others are not -- we assume that the covariates act **proportionally** on the hazard irrespective of time.

## Cox Proportional Hazards

- The model is estimated by maximizing a **partial likelihood**
  - We observe event time $Y_i$ - assume no ties

- The individual partial likelihood is

  $$L_i(\beta) = \frac{\exp(X_i^{\prime}\beta)}{\sum_{k; Y_k \ge Y_i} \exp(X_k^{\prime}\beta)}$$


- **Intuitively**: The denominator contains the hazards of all observations in the **risk set** at the event time $Y_i$

- Then the full likelihood is just the product over all the individual partial likelihoods

  $$L(\beta) = \prod_{i: C_i = 0}  \frac{\exp(X_i^{\prime}\beta)}{\sum_{k; Y_k \ge Y_i} \exp(X_k^{\prime}\beta)}$$

## Cox and conditional logit

::: {.incremental}

- The individual partial likelihood should look familiar - it has exactly the form of the **conditional logit**

  $$L_i(\beta) = \frac{\exp(X_i^{\prime}\beta)}{\sum_{k; Y_k \ge Y_i} \exp(X_k^{\prime}\beta)}$$

- At each event time we are asking: given that **one** unit in the risk set failed, which one was it?
  - The "choice set" is the risk set $\{k: Y_k \ge Y_i\}$, and the "chosen alternative" is the unit that failed.

:::

## Application

- What does the Cox fit look like

```{r, echo=T}
# sex = 2 is a dummy variable for female
cox_model_sex <- coxph(Surv(time, status) ~ I(sex==2), data=lung)
print(cox_model_sex)
```

- Note that we **don't get** an estimate of the hazard because the baseline is left unspecified
  - If we want to plot survival curves, we'll need some estimate of $h_0$ using another method (like Kaplan-Meier)

## Application

- Estimate the baseline cumulative hazard non-parametrically (the Breslow estimator) and get the Cox-predicted survival curves for men and women

```{r, echo=T}
cox_surv_sex <- survfit(cox_model_sex, newdata = data.frame(sex = c(1, 2)))
```

```{r, fig.width=8, fig.height=5, fig.align='center', warning=F}
plot(cox_surv_sex, col=c("indianred", "dodgerblue"), xlab="Survival Time", conf.int=T, ylab="Survival Rate", mark.time=T)
```

## Next week

::: {.incremental}

- **Introduction to Bayesian inference**
  - What happens if we assume parameters are not fixed but carry uncertainty?
  - The **posterior distribution** $p(\theta | \mathbf{Y})$ as a target of inference
  - The **prior distribution** and how it shapes the posterior.
- **Midterm Exam**
  
:::

## References {.refs-slide}

::: {#refs}
:::
