---
title: "PS 818: Problem Set 3"
author: "YOUR NAME"
date: "October 7, 2026"
format:
  html:
    toc: true
    df-print: paged
    embed-resources: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Useful packages
library(tidyverse)
library(truncnorm)
```

> **Assigned:** Wednesday, October 7, 2026. **Due:** Wednesday, October 21, 2026.
>
> Submit your solutions on **Gradescope**. Upload *both* the rendered `.html` file (saved as `Yourlastname_Yourfirstinitial_pset3.html`) and the `.qmd` source file that produced it (saved as `Yourlastname_Yourfirstinitial_pset3.qmd`). We should be able to run your code without error messages. In order to receive credit, homework submissions must be substantially started and all work must be shown.
>
> I recommend editing this `.qmd` file directly and rendering it to `.html` when you are done. Problem sets are graded on a (+/✓/-) scale.
>
> You are strongly encouraged to discuss the problems with other students, in person or on the course [Slack](https://uw-mpolscimethods.slack.com/), but each student must submit their own write-up and code. See the [Syllabus](https://www.antonstrezhnev.com/ps818/syllabus.html) for the full collaboration policy and for the course policy on generative large language models.

# Problem 1

This problem will have you implementing a Bayesian version of the classic probit regression model. The dataset comes from Barrilleaux and Rainey (2014), which studies the decisions of U.S. state governors to accept or reject the Affordable Care Act's Medicaid expansion. The full citation is:

> Barrilleaux, Charles, and Carlisle Rainey. "The politics of need: Examining governors’ decisions to oppose the “Obamacare” Medicaid expansion." *State Politics & Policy Quarterly* 14, no. 4 (2014): 437-460.

The code below will load in the dataset.

```{r, warning=F, message=F}
aca <- read_csv("data/politics_and_need.csv")
```

## Part A

We will need to first construct our outcome and covariates of interest. Create a binary indicator variable, `gov_opposes`, for whether a state's governor **opposes** Medicaid expansion by transforming the `gov_position` variable (treat both "Supports" and "Weighing Options" as not opposing). Your predictors/regressors will be:

- `gop_governor` - An indicator for whether the governor is a Republican (transform from `gov_party`)
- `percent_favorable_aca` - Percent of residents favorable to the ACA
- `gop_legislature` - An indicator for whether both houses of the state legislature are controlled by the Republican party (transform from `sen_party` and `house_party`)
- `percent_uninsured` - Percentage of residents who are uninsured
- `bal2012` - Fiscal health (state year-end reserves as a share of total spending)
- `multiplier` - State Medicaid multiplier
- `percent_nonwhite` - Percentage of residents who are non-white (transform from `percent_white`)
- `percent_metropolitan` - Percentage of residents who live in metropolitan areas

Some of the above variables do not exist in the data frame. Generate them by transforming the relevant variables as indicated.

Rescale each of the continuous regressors so that it has mean zero and standard deviation $.5$ (de-mean each variable and divide it by 2 times its sample standard deviation). Rescale each binary regressor so that it has mean zero.

## Part B

Next, let's write down a model. We'll assume that the indicator $Y_i \in \{0, 1\}$ of whether the governor opposes ACA Medicaid expansion has the following data-generating process:

$$Y_i = \mathbf{1}(Y_i^* > 0)$$

where $\mathbf{1}()$ denotes the indicator function. That is, when $Y_i^*$ is greater than $0$, $Y_i = 1$ and when $Y_i^*$ is less than or equal to $0$, $Y_i = 0$.

$$Y_i^* \sim \text{Normal}(X_i^{\prime}\beta, 1)$$

and

$$\beta \sim \text{Normal}(b_0, B_0^{-1})$$

where $b_0$ is a vector of hyperparameters denoting the prior mean of $\beta$ and $B_0$ is a matrix of hyperparameters denoting the inverse of the prior variance-covariance matrix of $\beta$.

Start by classifying each of the variables in the DGP. What is the **observed data**? What is a **latent variable** and what is a **known constant**?

## Part C

Our goal is to do inference on the posterior distribution

$$f(\beta \mid \mathbf{Y}, \mathbf{X}) \propto f(\mathbf{Y} \mid \beta, \mathbf{X})f(\beta \mid \mathbf{X}) \equiv f(\mathbf{Y} \mid \beta, \mathbf{X})f(\beta)$$

However, this is difficult. Instead, we simulate from the joint posterior that includes the $\mathbf{Y}^* = \{Y_1^*, Y_2^*, Y_3^*, \dotsc, Y_N^*\}$:

$$f(\beta, \mathbf{Y}^* \mid \mathbf{Y}, \mathbf{X})$$

We will then marginalize over this joint distribution by keeping the draws of $\beta$.

Use Bayes' rule to write this posterior distribution as proportional to a **likelihood** multiplied by a **prior**. Factor the density as much as possible and use the independencies and conditional independencies implied by the model to eliminate terms from the conditioning set of some of the densities. You should have three main component densities: one related to the prior on $\beta$ and two likelihood terms.

## Part D

What is the conditional distribution of $Y_i^* \mid Y_i, \beta, X_i$?

Hint: What do we already know is the conditional distribution of $Y_i^* \mid \beta, X_i$? What information does adding $Y_i$ provide?

## Part E

What is the form of the conditional distribution of $\beta \mid \mathbf{Y}^*, \mathbf{Y}, \mathbf{X}$? (Just the form/type of the distribution. You don't need to derive its mean and variance, though this is a standard result.)

Hint 1: What can we get rid of on the right-hand side of the conditioning bar (using conditional independence)?

Hint 2: Once we've gotten rid of that variable, what familiar posterior distribution from a different model does this remind you of?

## Part F

Let's estimate the probit model to predict the probability of opposing Medicaid expansion conditional on the covariates from Part A. Include the covariates additively, omitting any interactions. Use `model.matrix()` to generate the regression matrix $\mathbf{X}$.

Implement a Gibbs sampler to obtain draws from the joint posterior distribution. Use a fairly diffuse prior for $\beta$ that sets $b_0 = 0$ and $B_0 = \frac{1}{9} \mathbf{I}$ where $\mathbf{I}$ is the identity matrix.

Pick some reasonable starting values for $\beta$ and start by sampling from the conditional distribution of $Y_i^*$ to generate "starting" values of $Y_i^*$. Iterate between sampling from the conditional distribution of $\beta$ and the conditional distribution of the $Y_i^*$. You only need to store the draws of $\beta$; there is no need to keep the draws of $\mathbf{Y}^*$ from every iteration.

Set the starting seed to `53703` and run the sampler for `2000` iterations after discarding the first `500` as a "burn-in" (so you should run it for 2,500 iterations in total). With only 50 states the full chain should take a few seconds at most, but it is still good practice to test your code with a handful of iterations first.

**Hint 1**: You will find the `truncnorm` R package useful.

**Hint 2**: Using well-known results, the distribution from Part E has variance-covariance matrix

$$\mathbf{V} = (B_0 + \mathbf{X}^{\prime}\mathbf{X})^{-1}$$

and mean

$$\mathbf{M} = \mathbf{V}(B_0b_0 + \mathbf{X}^{\prime}\mathbf{Y}^*)$$

## Part G

Make some traceplots of your coefficients. Does it look like the Gibbs sampling chain has sufficiently "mixed"?

## Part H

Plot the posterior distribution of the coefficient on "GOP Governor". Obtain a posterior mean estimate along with a 95% credible interval. Interpret your results substantively. Compare the results to the estimates from a conventional probit regression estimated via MLE (use `glm()` to fit it). Why do we need to use a Bayesian approach for this particular regression setting?

## Part I

Now implement the model above in Stan without the data augmentation step (define $Y_i$ conventionally using the normal CDF `Phi`). Instead of using a normal prior, use a Cauchy prior with center 0 and scale parameter 6.25 for the intercept and a Cauchy prior with center parameter 0 and scale parameter 1.6 for the rest of the coefficients. Set the seed to $53703$. Obtain the posterior mean and 95% credible interval for the coefficient on "GOP Governor" and compare the results to your analysis above. How does changing the prior change the results?

**Hint**: Install RStan following the [RStan Getting Started guide](https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started), using the version from the Stan r-universe rather than CRAN: `install.packages("rstan", repos = c("https://stan-dev.r-universe.dev", getOption("repos")))`. Make sure your C++ toolchain is configured first (see the guide).

```{r}
library(rstan)
```
